### 2024-05-21 paper/main.tex
begin paper

\documentclass[conference]{IEEEtran}
\IEEEoverridecommandlockouts
% The preceding line is only needed to identify funding in the first footnote. If that is unneeded, please comment it out.
\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}
\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
\begin{document}

\title{Predicting Diamond Prices*\\
{\footnotesize \textsuperscript{*}ECS 171 Final Project - \textbf{Group 21}}
\thanks{Final Project Report for ECS 171 - Group 21}
}

\author{\IEEEauthorblockN{Aditya Mittal}
\IEEEauthorblockA{\textit{Department of Statistics} \\
\textit{University of California, Davis}\\
Davis, United States \\
adimittal@ucdavis.edu}
\and
\IEEEauthorblockN{2\textsuperscript{nd} Given Name Surname}
\IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
\textit{name of organization (of Aff.)}\\
City, Country \\
email address or ORCID}
\and
\IEEEauthorblockN{3\textsuperscript{rd} Given Name Surname}
\IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
\textit{name of organization (of Aff.)}\\
City, Country \\
email address or ORCID}
\and
\IEEEauthorblockN{4\textsuperscript{th} Given Name Surname}
\IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
\textit{name of organization (of Aff.)}\\
City, Country \\
email address or ORCID}
\and
\IEEEauthorblockN{5\textsuperscript{th} Given Name Surname}
\IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
\textit{name of organization (of Aff.)}\\
City, Country \\
email address or ORCID}
}

\maketitle

\begin{abstract}
This document is a model and instructions for \LaTeX.
This and the IEEEtran.cls file define the components of your paper [title, text, heads, etc.]. *CRITICAL: Do Not Use Symbols, Special Characters, Footnotes, 
or Math in Paper Title or Abstract.
\end{abstract}

\begin{IEEEkeywords}
component, formatting, style, styling, insert
\end{IEEEkeywords}

\section{Introduction}
This document is a model and instructions for \LaTeX.
Please observe the conference page limits.

\section{Literature Review}

\section{Dataset Description and Exploratory Data Analysis}

\section{Proposed methodology}

\section{Experimental results and evaluation}

\section{Conclusion}

\section*{Acknowledgment}

\section*{References}

\begin{thebibliography}{00}
\bibitem{b1} G. Eason, B. Noble, and I. N. Sneddon, ``On certain integrals of Lipschitz-Hankel type involving products of Bessel functions,'' Phil. Trans. Roy. Soc. London, vol. A247, pp. 529--551, April 1955.
\bibitem{b2} J. Clerk Maxwell, A Treatise on Electricity and Magnetism, 3rd ed., vol. 2. Oxford: Clarendon, 1892, pp.68--73.
\bibitem{b3} I. S. Jacobs and C. P. Bean, ``Fine particles, thin films and exchange anisotropy,'' in Magnetism, vol. III, G. T. Rado and H. Suhl, Eds. New York: Academic, 1963, pp. 271--350.
\bibitem{b4} K. Elissa, ``Title of paper if known,'' unpublished.
\bibitem{b5} R. Nicole, ``Title of paper with only first word capitalized,'' J. Name Stand. Abbrev., in press.
\bibitem{b6} Y. Yorozu, M. Hirano, K. Oka, and Y. Tagawa, ``Electron spectroscopy studies on magneto-optical media and plastic substrate interface,'' IEEE Transl. J. Magn. Japan, vol. 2, pp. 740--741, August 1987 [Digests 9th Annual Conf. Magnetics Japan, p. 301, 1982].
\bibitem{b7} M. Young, The Technical Writer's Handbook. Mill Valley, CA: University Science, 1989.
\end{thebibliography}
\vspace{12pt}

\end{document}

### 2024-06-01 code/exploratory_data_analysis.py
Update exploratory_data_analysis.py

@@ -1,1 +1,149 @@
+# import necessary modules
+import pandas as pd
+import matplotlib.pyplot as plt
+import seaborn as sns
+import numpy as np
+from sklearn.linear_model import LinearRegression
+from sklearn.preprocessing import PolynomialFeatures
+from sklearn.preprocessing import StandardScaler
+from sklearn.model_selection import train_test_split
+from sklearn.metrics import mean_squared_error, r2_score
+from sklearn.preprocessing import MinMaxScaler
+import tensorflow as tf
+from tensorflow.keras.models import Sequential
+from tensorflow.keras.layers import Dense, Dropout
+from tensorflow.keras import regularizers
+from sklearn.linear_model import Ridge, Lasso
+from sklearn.model_selection import train_test_split, GridSearchCV, KFold
+from sklearn.preprocessing import StandardScaler
+from tensorflow.keras.models import Sequential
+from tensorflow.keras.layers import Dense, Dropout
+from tensorflow.keras.optimizers.legacy import SGD 
+from tensorflow.keras.wrappers.scikit_learn import KerasRegressor
+from tensorflow.keras.regularizers import l1, l2
 
+df = pd.read_csv("~/Desktop/Year_three/Spring_2024/ECS171/ECS171_Final/diamonds.csv") 
+df = df.iloc[:, 1:]  # remove index column           
+df.head()
+
+### EXPLORATORY DATA ANALYSIS
+### Plot distribution of continous features 
+features_to_plot = ['carat', 'depth', 'table', 'x','y','z']
+plt.figure(figsize=(8, 6))
+for i, feature in enumerate(features_to_plot, 1):
+    plt.subplot(2, 3, i)
+    sns.histplot(df[feature], kde=True)
+    plt.title(feature)
+
+plt.tight_layout()
+plt.show()
+print("\nmean:\n{} , \nmedian:\n{} , \nstd:\n {}".format(df[features_to_plot].mean() , 
+                                                         df[features_to_plot].median(), 
+                                                         df[features_to_plot].std()))
+
+### Plot Histogram frequency of categorical variables
+categorical_vars = ["cut", "color", "clarity"]
[…]

### 2024-06-01 code/exploratory_data_analysis.py
Delete code/exploratory_data_analysis.py

@@ -1,149 +1,0 @@
-# import necessary modules
-import pandas as pd
-import matplotlib.pyplot as plt
-import seaborn as sns
-import numpy as np
-from sklearn.linear_model import LinearRegression
-from sklearn.preprocessing import PolynomialFeatures
-from sklearn.preprocessing import StandardScaler
-from sklearn.model_selection import train_test_split
-from sklearn.metrics import mean_squared_error, r2_score
-from sklearn.preprocessing import MinMaxScaler
-import tensorflow as tf
-from tensorflow.keras.models import Sequential
-from tensorflow.keras.layers import Dense, Dropout
-from tensorflow.keras import regularizers
-from sklearn.linear_model import Ridge, Lasso
-from sklearn.model_selection import train_test_split, GridSearchCV, KFold
-from sklearn.preprocessing import StandardScaler
-from tensorflow.keras.models import Sequential
-from tensorflow.keras.layers import Dense, Dropout
-from tensorflow.keras.optimizers.legacy import SGD 
-from tensorflow.keras.wrappers.scikit_learn import KerasRegressor
-from tensorflow.keras.regularizers import l1, l2
-
-df = pd.read_csv("~/Desktop/Year_three/Spring_2024/ECS171/ECS171_Final/diamonds.csv") 
-df = df.iloc[:, 1:]  # remove index column           
-df.head()
-
-### EXPLORATORY DATA ANALYSIS
-### Plot distribution of continous features 
-features_to_plot = ['carat', 'depth', 'table', 'x','y','z']
-plt.figure(figsize=(8, 6))
-for i, feature in enumerate(features_to_plot, 1):
-    plt.subplot(2, 3, i)
-    sns.histplot(df[feature], kde=True)
-    plt.title(feature)
-
-plt.tight_layout()
-plt.show()
-print("\nmean:\n{} , \nmedian:\n{} , \nstd:\n {}".format(df[features_to_plot].mean() , 
-                                                         df[features_to_plot].median(), 
-                                                         df[features_to_plot].std()))
-
-### Plot Histogram frequency of categorical variables
-categorical_vars = ["cut", "color", "clarity"]
[…]

### 2024-06-01 paper/main.tex
Update main.tex

@@
 \usepackage{textcomp}
+\usepackage{natbib}
+\usepackage{float}
 \usepackage{xcolor}
@@
 
-\title{Predicting Diamond Prices*\\
-{\footnotesize \textsuperscript{*}ECS 171 Final Project - \textbf{Group 21}}
+\title{Machine Learning Approaches for Accurate Diamond Price Prediction\\
+% {\footnotesize ECS-171 Final Project: \textbf{Group 21}}
 \thanks{Final Project Report for ECS 171 - Group 21}
@@
 
-\author{\IEEEauthorblockN{Aditya Mittal}
+\author{
+\IEEEauthorblockN{Aditya Mittal}
 \IEEEauthorblockA{\textit{Department of Statistics} \\
@@
 \and
-\IEEEauthorblockN{2\textsuperscript{nd} Given Name Surname}
-\IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
-\textit{name of organization (of Aff.)}\\
-City, Country \\
-email address or ORCID}
+\IEEEauthorblockN{Andrew Yeow}
+\IEEEauthorblockA{\textit{Department of Computer Science} \\
+\textit{University of California, Davis}\\
+Davis, United States \\
+XXX}
 \and
-\IEEEauthorblockN{3\textsuperscript{rd} Given Name Surname}
-\IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
-\textit{name of organization (of Aff.)}\\
-City, Country \\
-email address or ORCID}
+\IEEEauthorblockN{Yifan Cui}
+\IEEEauthorblockA{\textit{Department of Computer Science} \\
+\textit{University of California, Davis}\\
+Davis, United States \\
+XXX}
 \and
-\IEEEauthorblockN{4\textsuperscript{th} Given Name Surname}
-\IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
-\textit{name of organization (of Aff.)}\\
-City, Country \\
-email address or ORCID}
+\IEEEauthorblockN{Nandini Koladi}
+\IEEEauthorblockA{\textit{Department of Computer Science} \\
+\textit{University of California, Davis}\\
+Davis, United States \\
+XXX}
 \and
-\IEEEauthorblockN{5\textsuperscript{th} Given Name Surname}
-\IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
-\textit{name of organization (of Aff.)}\\
-City, Country \\
-email address or ORCID}
+\IEEEauthorblockN{Eshan Toshniwal}
+\IEEEauthorblockA{\textit{Department of Computer Science} \\
+\textit{University of California, Davis}\\
+Davis, United States \\
+XXX}
 }
@@
 \begin{abstract}
-This document is a model and instructions for \LaTeX.
-This and the IEEEtran.cls file define the components of your paper [title, text, heads, etc.]. *CRITICAL: Do Not Use Symbols, Special Characters, Footnotes, 
-or Math in Paper Title or Abstract.
+In 2023, the U.S.
+jewelry market was valued at 73.32 billion USD, with diamonds comprising a significant share.
+Predicting diamond prices accurately is essential for jewelers to make informed pricing decisions.
+This paper explores various machine learning algorithms, including linear regression, Random Forests, XGBoost, and neural networks, to predict diamond prices based on characteristics such as carat, cut, color, and clarity.
+Through exploratory data analysis, we identify key features influencing diamond prices and apply these insights to improve model performance.
+Our methodology includes rigorous model evaluation using metrics like Mean Squared Error (MSE), Mean Absolute Error (MAE), and R-squared (R²).
+By developing a robust predictive model, we aim to enhance pricing precision, benefiting both businesses and consumers in the diamond market.
 \end{abstract}
@@
 \begin{IEEEkeywords}
-component, formatting, style, styling, insert
+Diamond pricing, machine learning, predictive modeling, neural network, decision trees
 \end{IEEEkeywords}
@@
 \section{Introduction}
-This document is a model and instructions for \LaTeX.
-Please observe the conference page limits.
+
+In 2023, the jewelry market in the United States was estimated to be worth 73.32 billion USD, with projections for 2024 reaching 96.61 billion USD \cite{a3}.
+Clearly, the market for jewelry is quite significant and continues to display steady growth.
+In particular, \emph{diamonds} account for a significant portion of this market, covering an estimated 50-60\% market share as compared to other precious stones such as gold, silver, etc \cite{a4}.
+Consequently, it becomes important for jewelers to adopt accurate methods to predict prices of these diamonds to better serve their consumers.
+In fact, businesses have began educating their sellers on essential qualities such as cut, color, carat weight, and clarity to make informed pricing decisions.
+As such, ensuring accurate price predictions becomes very important as it has significant financial implications on both businesses and consumers alike.
+
+Given the importance of this task, it becomes crucial to develop methods that enable accurate prediction of prices for these diamonds.
+Machine learning, with its ability to learn patterns in data and make predictions, has become extremely popular in recent years across a wide range of applications \cite{a1}.
+Recent advancements in machine learning have introduced a range of algorithms, including linear regression, K-Nearest Neighbors, Random Forests, and the increasingly popular neural networks \cite{c1}.
+In this paper, we explore several different machine learning algorithms to accurately predict diamond prices based on several characteristics, including those mentioned above.
+The application of these predictions is widespread, as it can aid jewelers and marketers who aim to provide accurate pricing for consumers in the jewelry market.
+
+Our dataset contains several essential features for diamond price prediction, such as the aforementioned carat, cut, color, clarity, etc \cite{a2}.
+Through exploratory data analysis (EDA), we aim to understand how these features relate to diamond prices.
+This analysis will help us answer questions about which features most significantly impact diamond prices and how we can leverage this information to create predictive models.
+Thus, our goal is to develop an accurate model for predicting diamond prices, enabling jewelers to make precise pricing decisions and enhancing the consumer purchasing experience.
+This paper details our methodology, including EDA and the application of various machine learning algorithms, to achieve this objective.
+
+The outline of this paper is as follows: Section 2 discusses the historical perspective of machine learning and relevant algorithms as part of the literature review.
+Section 3 presents data description and exploratory data analysis.
+Sections 4 and 5 discuss the proposed methodology and results.
+Finally, the paper concludes with a discussion and future perspective in Section 6.
 
@@
 
+Machine learning is a rapidly developing field with a diverse range of algorithms tailored for various tasks.
+This section details important concepts surrounding the ML algorithms we will use.
+In particular, we discuss the historical context \& purpose of regression models, decision trees, and neural networks.
+
+\subsection{Linear Regression}
+
+The most basic prediction model we explore is linear regression.
+This model fits a straight line between the response variable and its predictors, making it extremely simple and interpretable in terms of estimating the regression coefficients \cite{b10}.
+One major benefit of linear regression, compared to more complex models, is its efficiency and ease of interpretation, especially when the relationship between the dependent variable $Y$ and the independent variable(s) $X$ is approximately linear.
+For a single predictor $X$, the general form of a linear regression model is:
+
+\[
+    Y = \beta_0 + \beta_1 X + \epsilon
+\]
+
+Here, $\beta_0$ represents the intercept term, $\beta_1$ is the coefficient for X, and $\epsilon_1 \sim \mathcal{N}(0,\sigma_1^2)$.
+Note that $\epsilon_1$ is independent of the other predictor terms.
+Linear regression relies on several key assumptions, including linearity, independence, homoscedasticity, and normality of residuals \cite{b8}.
+The performance of a linear regression model can be evaluated using metrics such as $R^2$ and Mean Squared Error (MSE).
+Linear regression can be extended to more complex models, such as multiple linear regression, polynomial regression, and regularization methods (e.g., Ridge and Lasso regression) to handle more complex relationships and preventing overfitting.
+
+A crucial part of linear regression is the minimization of the Residual Sum of Squares (RSS). The RSS measures the discrepancy between the observed data and the values predicted by the model:
+
+\[
+    \text{RSS} = \sum_{i=1}^n (y_i - \hat{y}_i)^2
+\]
+
+Minimizing the RSS through Ordinary Least Squares (OLS) provides the best-fitting line by ensuring the smallest possible error in predicting the response variable from the predictors.
+For further details on the mathematical foundations of OLS and RSS minimization, refer to Seber and Lee's comprehensive discussion on linear regression \cite{b9}.
+Linear regression has been extensively studied and applied across various fields \cite{b1, b2}.
+Additionally, Kutner et al.
+provide detailed discussions on the statistical assumptions, limitations, and diagnostics of linear regression models \cite{b8}.
+
+\subsection{Decision Trees}
+
+The second major algorithm we review is decision trees.
+Decision trees are powerful tools for classification and regression tasks, especially with tabular data.
+These algorithms split the data into subsets based on the value of input features, creating a tree-like model of decisions.
+Decision trees can be prone to overfitting; however, ensemble methods such as Random Forests (Breiman, 2001) and Gradient Boosting Machines like XGBoost (Chen \& Guestrin, 2016) mitigate this issue by combining multiple trees to improve performance and robustness.
+Random forests, introduced by Breiman (2001), create a "forest" of multiple decision trees and aggregate their predictions.
+This method reduces variance and improves generalization by averaging multiple models .
+XGBoost, a popular implementation of gradient boosting, has been shown to deliver state-of-the-art performance in many machine learning competitions due to its efficiency and scalability.
+
+Despite being recently popularized by the rise of AI applications, neural networks are actually a historical concept developed in the early 1980s.
+It resurgence was sparked by the rise of deep learning, with advancements in computational power to handle larger datasets.
+
+\subsection{Neural Networks}
+
+The foundation of modern neural networks was laid by the introduction of backpropagation by Rumelhart, Hinton, and Williams (1986) .
+Recently, deep learning has revolutionized fields such as image recognition, natural language processing, and game playing.
+LeCun, Bengio, and Hinton (2015) provide a seminal review on the principles and advancements in deep learning.
+A neural network contains many hyperparameters, and there are several elements to finding the best set of such values.
+Grid search is a systematic approach for hyperparameter tuning, where a predefined set of hyperparameters is evaluated exhaustively to identify the optimal configuration for a machine learning model.
+Bergstra and Bengio (2012) introduced grid search as a fundamental method for hyperparameter optimization.
+They highlighted its importance in efficiently exploring the hyperparameter space and provided insights into its limitations, such as scalability and computational cost.
+For a comprehensive discussion on hyperparameter optimization techniques, see the review by Loshchilov and Hutter (2019).
+
+The Adam optimizer, short for Adaptive Moment Estimation, is a widely used optimization algorithm for training deep neural networks.
+It combines the advantages of both adaptive learning rate methods and momentum-based optimization techniques, offering fast convergence and robustness to noisy gradients.
+Kingma and Ba (2014) introduced the Adam optimizer, presenting it as an extension of the stochastic gradient descent with adaptive learning rates.
+They demonstrated its effectiveness in training deep learning models across various tasks, including image classification, natural language processing, and reinforcement learning.
+Adam maintains adaptive learning rates for each parameter by computing individual learning rates based on the first and second moments of the gradients.
+This adaptive behavior allows Adam to automatically adjust the learning rate during training, leading to improved convergence and generalization performance.
+
+In conclusion, grid search and the Adam optimizer are essential components of the machine learning and deep learning toolkit.
+While grid search provides a systematic approach for hyperparameter tuning, Adam offers an efficient optimization algorithm for training deep neural networks.
+Understanding their strengths, limitations, and best practices is crucial for effectively applying these techniques in practical scenarios.
+
 \section{Dataset Description and Exploratory Data Analysis}
 
+The \emph{diamond prices} dataset contains detailed information about 54,000 diamonds, including their prices and various characteristics \cite{a2}.
+The predictor variables in this dataset are as follows: i) \textbf{Carat}: Represents the weight of the diamond, ranging from 0.2 to 5.01 carats.
+ii) \textbf{Cut Quality}: Categorized as Fair, Good, Very Good, Premium, or Ideal, indicating the craftsmanship involved in shaping the diamond.
+iii) \textbf{Color}: Graded from J (worst) to D (best), reflecting the hue and saturation of the stone.
+iv) \textbf{Clarity}: Measures how clear the diamond is, with grades ranging from I1 (worst), SI2, SI1, VS2, VS1, VVS2, VVS1, to IF (best).
+v) \textbf{Dimensions}: Captured by the length (x), width (y), and depth (z) in millimeters, with ranges of 0–10.74 mm, 0–58.9 mm, and 0–31.8 mm, respectively.
+vi) \textbf{Depth Percentage}: Ranges from 43\% to 79\%.
+vii) \textbf{Table}: Represents the width of the top of the diamond relative to its widest point, ranging from 43\% to 95\%.
+The dataset does not contain any missing values, so we do not need to address missing data in the exploratory data analysis step.
+
+\subsection{Exploratory Data Analysis}
+
+To explore the dataset further and understand the relationships between these features and diamond prices, we will conduct exploratory data analysis (EDA).
+First, we visualize the distribution of each continuous feature, such as carat, length (x), width (y), depth (z), depth percentage, and table, to understand their spread and central tendencies.
+A table will highlight the mean, median, and standard deviation statistics of each feature, providing insights into their individual magnitudes and spread.
+Second, pairwise scatter plots display the relationship between each continuous feature and diamond price, helping to identify potential trends and correlations.
+Third, the correlation heat-map presents Pearson correlation relationships between continuous features, helping identify features most correlated with price.
+Finally, we create histograms of categorical features like cut, color, and clarity to understand their distributions.
+These visualizations provide insights into the structure of the data, highlight any anomalies, and help with subsequent modeling efforts to predict diamond prices effectively.
+
+\begin{figure}[H]
+    \centering
+    \includegraphics[width=0.8\linewidth]{Figure_1.png}
+    \caption{Distribution Plot of Continuous Features. Reveals trends in data such as skewness, location of outliers, etc.}
+    \label{fig:image_label}
+\end{figure}
+
+The distribution of continuous predictors highlight potential skewness in our dataset.
+For instance, the predictor \emph{carat} suggests potential skewness to the right – with many upper outliers – as evidenced by its distribution plot.
+This trend holds true for features \emph{y} and \emph{z} as well.
+These plots suggest that our dataset may contain a significant number of upper outliers, which we will consider removing in the data pre-processing stage.
+To further gain insight on the distribution of our features, the table below summarizes the mean, median, and standard deviation of the continuous features:
+
+\begin{table}[H]
+    \centering
+    \caption{Mean, Median, Standard Dev. of Continuous Features}
+    \label{tab:example_table}
+    \begin{tabular}{|c|c|c|c|}
+        \hline
+        Feature & Mean & Median & Standard Dev. \\
+        \hline
+        Carat & 0.797 & 0.70 & 0.474 \\
+        \hline
+        Depth & 61.749 & 61.80  & 1.432 \\
+        \hline
+        Table & 57.457 & 57.00 & 2.234 \\
+        \hline
+        X & 5.731 & 5.70 & 1.122 \\
+        \hline
+        Y & 5.734 & 5.71 & 1.142 \\
+        \hline
+        Z & 3.538 & 3.53 & 0.705 \\
+        \hline
+    \end{tabular}
+\end{table}
+
+From our table, we can observe that the mean and median values for most features are relatively close, indicating that the data distributions may not actually be as heavily skewed for most variables.
+However, the standard deviation values highlight the variability within each feature, such as carat and dimensions (x, y, z), display considerable spread.
+This variability and potential skewness, especially in the carat feature from the plots, could influence our predictive algorithms.
+Next, we will create pairwise scatter-plots to identify trends between diamond prices and the features.
+
+\begin{figure}[H]
+    \centering
+    \includegraphics[width=0.8\linewidth]{Figure_4.png} % Replace "example-image" with the filename of your image
+    \caption{Pair-wise Scatterplots for Each Feature vs. Diamond Price}
+    \label{fig:image_label}
+\end{figure}
+
+From the scatterplots of each feature against the response variable, we observe a quadratic relationship between the features, carat, x, y, and z, with respect to the the response variable price.
+This suggests that a polynomial regression model might be suitable for our dataset in addition to a linear regression model.
+Thus, we will evaluate the polynomial regression model's performance on the dataset compared to traditional linear regression.
+Next, we will also analyze the correlations between these continuous features and the target variable, price, to determine the most influential predictors.
+This will allow us to identify the most relevant features and also potential multicollinearity and interactions.
+
+\begin{figure}[H]
+    \centering
+    \includegraphics[width=0.8\linewidth]{Figure_3.png} % Replace "example-image" with the filename of your image
+    \caption{Correlation Heat-Map for Continous Features vs Diamond Price}
+    \label{fig:image_label}
+\end{figure}
+
+Based on the results of our correlation heatmap, the feature \emph{Carat} appears to be highly correlated with price, with a correlation coefficient of 0.92.
+Similarly, it is correlated with dimension features \emph{x}, \emph{y}, and \emph{z}.
+This makes sense, as carat weight would directly relate to diamond size.
+Thus, this plot suggests that Carat (and potentially dimensions) are the strongest predictors for diamond price.
+We showcase data preprocessing methods for outlier removal using carat below.
+On the other hand, both depth percentage and table have almost zero correlation, indicating that they do not have much association with diamond prices.
+Thus, this heatmap is insightful in identifying the key predictors for diamond prices and guiding our feature selection process.
+Next, we will examine histograms for the distribution of categorical features cut, clarity, and color:
+
+\begin{figure}[H]
+    \centering
+    \includegraphics[width=0.8\linewidth]{figure_5.png} % Replace "example-image" with the filename of your image
+    \caption{Historgram for the distribution of categorical features: Cut, Clarity, and Color }
+    \label{fig:image_label}
+\end{figure}
+
+The histogram of the categorical variables displays their frequencies.
+Upon initial analysis of the plot, we observe that the cut category "Ideal" appears most frequently, followed by "Premium" and "Very Good," while the "Fair" cut appears least frequently.
+The distribution of color appears to be well spread, with all colors represented.
+Regarding clarity, categories such as "I1" and "IF" have relatively few observations compared to others.
+With this, our exploratory data analysis step has provided valuable insights into the structure of the dataset and identify patterns among the features.
+We will now proceed with data pre-processing before implementing our machine learning algorithms.
+
+\subsection{Data Pre-Processing}
+
+Before implementing the machine learning algorithms, we conducted several pre-processing steps on our dataset.
+First, considering the ordinality present in each of our categorical features, we implemented label encoding to convert these features into numeric representations.
+The mappings for each categorical feature were defined as follows:
+
+\begin{table}[htbp]
+    \centering
+    \caption{Label Encodings for Categorical Features}
+    \resizebox{\columnwidth}{!}{
+        \begin{tabular}{|l|l|}
+            \hline
+            \textbf{Feature} & \textbf{Encoding} \\
+            \hline
+            Cut & 'Fair': 0, 'Good': 1, 'Very Good': 2, 'Premium': 3, 'Ideal': 4 \\
+            Color & 'J': 0, 'I': 1, 'H': 2, 'G': 3, 'F': 4, 'E': 5, 'D': 6 \\
+            Clarity & 'I1': 0, 'SI2': 1, 'SI1': 2, 'VS2': 3, 'VS1': 4, 'VVS2': 5, 'VVS1': 6, 'IF': 7 \\
+            \hline
+        \end{tabular}
+    }
+    \label{tab:label_encodings}
+\end{table}
+
+Next, to standardize our data and balance the scale of our features, we applied z-score transformation.
+This transformation ensures that all our features have a mean of 0 and a standard deviation of 1, thereby balancing the scales among features.
+This is particularly beneficial for algorithms such as neural networks.
+Finally, we proceeded to remove outliers.
+Since carat was our most highly correlated predictor, we used it as the feature to identify and remove outliers using the 1.5 times the interquartile range (IQR) calculation.
+Approximately 3.9\% of the dataset was removed through this process, leaving us with 51,800 rows of data.
+The boxplots for each feature post outlier removal are shown below:
+
+\begin{figure}[H]
+    \centering
+    \includegraphics[width=0.8\linewidth]{boxplot.png} % Replace "example-image" with the filename of your image
+    \caption{Boxplots of Features Post-Outlier Removal}
+    \label{fig:boxplots}
+\end{figure}
+
+The boxplots indicate that outliers were removed for carat, while several other features such as depth and table still exhibit upper values.
+Since these features are not highly correlated with our response variable, we decided to keep them as is and proceed with this cleaned version of the dataset.
+
 \section{Proposed methodology}
+
+In this paper, we explore the predictive capabilities of several machine learning algorithms, ranging from linear regression and polynomial regression to decision trees and artificial neural networks.
+Our approach involves multiple steps to build and evaluate models that can accurately predict diamond prices based on various features.
+
+We begin with traditional linear regression to fit a straight-line model between diamond prices and the aforementioned features.
+This initial step helps us gain intuition about the relationship between the features and diamond prices using a simple model.
+Recognizing that the price exhibited a quadratic relationship with carat and diamond dimensions, we extend our analysis to polynomial regression by including squared terms of these features.
+This allows us to capture potential nonlinear relationships and assess whether polynomial regression outperforms traditional linear regression.
+
+For decision tree-based models, we employ two powerful algorithms: Random Forests and XGBoost.
+Random Forests aggregate the predictions of multiple decision trees to enhance accuracy and reduce overfitting.
+XGBoost, a gradient boosting algorithm, further refines predictions by optimizing the model iteratively to minimize errors.
+These models are well-suited for capturing complex interactions between features and are expected to perform robustly on our dataset.
+
+Finally, we explore the use of artificial neural networks (ANNs) for generating predictions.
+To optimize our ANN, we conduct a grid search for hyperparameter tuning, ensuring we select the best set of parameters for our model.
+Our neural network architecture consists of a 3-layer ANN with 64 neurons in the first hidden layer, using the ReLU activation function.
+The output layer also employs the ReLU activation function, as our task involves regression.
+
+To evaluate the performance of our models, we will compare their results using several metrics, including Mean Squared Error (MSE), Mean Absolute Error (MAE), and R-squared (R²).
+These metrics will help us determine the accuracy and reliability of each model on testing data.
+By comparing these measures, we aim to identify the most effective algorithm for predicting diamond prices, providing valuable insights for jewelers and marketers in the industry.
+
+Through this comprehensive methodology, we seek to develop an accurate and robust model for predicting diamond prices, leveraging the strengths of various machine learning techniques to achieve the best possible performance.
 
@@
 
-\section*{Acknowledgment}
+In this study, we investigated various machine learning algorithms to predict diamond prices based on key characteristics such as carat, cut, color, clarity, and dimensions.
+Our analysis demonstrated that advanced machine learning techniques, particularly Random Forests, XGBoost, and neural networks, significantly outperformed traditional linear regression models in terms of predictive accuracy.
+The exploratory data analysis revealed that carat is the most influential feature, with strong correlations to price and other dimensions.
 
-\section*{References}
+Our findings suggest that leveraging these advanced models can provide jewelers and consumers with more accurate pricing, enhancing decision-making and market transparency.
+Future work can explore integrating additional features, refining model hyperparameters, and applying these models to other datasets to generalize our findings further.
+
+By adopting robust machine learning methodologies, the diamond industry can benefit from more precise pricing strategies, ultimately leading to improved market efficiency and customer satisfaction.
 
 \begin{thebibliography}{00}
-\bibitem{b1} G.
-Eason, B.
-Noble, and I.
-N.
-Sneddon, ``On certain integrals of Lipschitz-Hankel type involving products of Bessel functions,'' Phil.
-Trans.
-Roy.
-Soc.
-London, vol.
-A247, pp.
-529--551, April 1955.
-\bibitem{b2} J. Clerk Maxwell, A Treatise on Electricity and Magnetism, 3rd ed., vol. 2. Oxford: Clarendon, 1892, pp.68--73.
-\bibitem{b3} I. S. Jacobs and C. P. Bean, ``Fine particles, thin films and exchange anisotropy,'' in Magnetism, vol. III, G. T. Rado and H. Suhl, Eds. New York: Academic, 1963, pp. 271--350.
-\bibitem{b4} K. Elissa, ``Title of paper if known,'' unpublished.
-\bibitem{b5} R. Nicole, ``Title of paper with only first word capitalized,'' J. Name Stand. Abbrev., in press.
-\bibitem{b6} Y.
-Yorozu, M.
-Hirano, K.
-Oka, and Y.
-Tagawa, ``Electron spectroscopy studies on magneto-optical media and plastic substrate interface,'' IEEE Transl.
-J.
-Magn.
-Japan, vol.
-2, pp.
-740--741, August 1987 [Digests 9th Annual Conf.
-Magnetics Japan, p.
-301, 1982].
-\bibitem{b7} M. Young, The Technical Writer's Handbook. Mill Valley, CA: University Science, 1989.
+\bibitem{a1} Kino, S., Hsu, Y.-T., Shiba, K., Chien, Y.-S., Mita, C., Kawachi, I., and Daoud, A.
+(2021).
+A scoping review on the use of machine learning in research on social determinants of health: Trends and research prospects.
+SSM - Population Health, 15, 100836.
+\bibitem{c1} Sarker, I.H. Machine Learning: Algorithms, Real-World Applications and Research Directions. SN COMPUT. SCI. 2, 160 (2021). https://doi.org/10.1007/s42979-021-00592-x
+\bibitem{a2} Agrawal, S. (2017a, May 25). Diamonds. Kaggle. https://www.kaggle.com/datasets/shivam2503/diamonds 
+\bibitem{a3} U.S.
+jewelry market size and share: Industry Report, 2030.
+U.S.
+Jewelry Market Size And Share | Industry Report, 2030.
+(n.d.).
+https://www.grandviewresearch.com/industry-analysis/us-jewelry-market-report
+\bibitem{a4} Diamond Jewelry Market Size \& Share Analysis Report, 2030. (n.d.). https://www.grandviewresearch.com/industry-analysis/diamond-jewelry-market-report 
+\bibitem{b1} Weisberg, S. (2005). Applied Linear Regression. Wiley.
+\bibitem{b2} Chatterjee, S., \& Hadi, A. S. (2015). Regression Analysis by Example. Wiley.
+\bibitem{b3} Breiman, L. (2001). Random Forests. Machine Learning, 45(1), 5-32.
+\bibitem{b4} Chen, T., \& Guestrin, C.
+(2016).
+XGBoost: A Scalable Tree Boosting System.
+In Proceedings of the 22nd ACM SIGKDD International Conference on Knowledge Discovery and Data Mining (pp.
+785-794).
+\bibitem{b5} Rumelhart, D. E., Hinton, G. E., \& Williams, R. J. (1986). Learning representations by back-propagating errors. Nature, 323(6088), 533-536.
+\bibitem{b6} LeCun, Y., Bengio, Y., \& Hinton, G. (2015). Deep learning. Nature, 521(7553), 436-444.
+\bibitem{b7}Vaswani, A., Shazeer, N., Parmar, N., Uszkoreit, J., Jones, L., Gomez, A.
+N., Kaiser, Ł., \& Polosukhin, I.
+(2017).
+Attention is all you need.
+In Advances in Neural Information Processing Systems (pp.
+5998-6008).
+\bibitem{b8}Michael Kutner, Christopher Nachtsheim, John Neter, and William Li. Applied Linear Statistical Models. 1974.
+\bibitem{b9}Seber, G. A. F., \& Lee, A. J. (2012). Linear Regression Analysis. Hoboken, NJ: John Wiley \& Sons. 
+\bibitem{b10} Wright, John T. "Linear Regression Analysis." British Medical Journal, vol. 310, no. 6977, 1995, pp. 1120-1124. BMJ Group
 \end{thebibliography}

### 2024-06-02 paper/main.tex
Update main.tex

@@
 \usepackage{xcolor}
+\bibliographystyle{plainnat}
 \def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
@@
 % {\footnotesize ECS-171 Final Project: \textbf{Group 21}}
-\thanks{Final Project Report for ECS 171 - Group 21}
+\thanks{Final Project Report ECS-171: \textbf{Group 21}}
 }
@@
 
-In 2023, the jewelry market in the United States was estimated to be worth 73.32 billion USD, with projections for 2024 reaching 96.61 billion USD \cite{a3}.
+In 2023, the jewelry market in the United States was estimated to be worth 73.32 billion USD, with projections for 2024 reaching 96.61 billion USD \cite{us}.
 Clearly, the market for jewelry is quite significant and continues to display steady growth.
-In particular, \emph{diamonds} account for a significant portion of this market, covering an estimated 50-60\% market share as compared to other precious stones such as gold, silver, etc \cite{a4}.
+In particular, \emph{diamonds} account for a significant portion of this market, covering an estimated 50-60\% market share as compared to some of the other precious stones such as gold, silver, etc \cite{diamond}.
 Consequently, it becomes important for jewelers to adopt accurate methods to predict prices of these diamonds to better serve their consumers.
@@
 Given the importance of this task, it becomes crucial to develop methods that enable accurate prediction of prices for these diamonds.
-Machine learning, with its ability to learn patterns in data and make predictions, has become extremely popular in recent years across a wide range of applications \cite{a1}.
-Recent advancements in machine learning have introduced a range of algorithms, including linear regression, K-Nearest Neighbors, Random Forests, and the increasingly popular neural networks \cite{c1}.
-In this paper, we explore several different machine learning algorithms to accurately predict diamond prices based on several characteristics, including those mentioned above.
-The application of these predictions is widespread, as it can aid jewelers and marketers who aim to provide accurate pricing for consumers in the jewelry market.
+Machine learning, with its ability to learn patterns in data and generate predictions, has become extremely popular in recent years across a wide range of applications \cite{kino}.
+Recent advancements in machine learning have introduced a range of algorithms, including linear regression, K-Nearest Neighbors, Random Forests, and the increasingly popular neural networks \cite{sarker}.
+In this paper, we explore several different machine learning algorithms to accurately predict diamond prices based on their given characteristics, including those mentioned above.
+The application of creating models to generate accurate predictions is widespread, as it can aid jewelers and marketers who aim to provide accurate pricing methods for consumers in the jewelry market.
 
-Our dataset contains several essential features for diamond price prediction, such as the aforementioned carat, cut, color, clarity, etc \cite{a2}.
-Through exploratory data analysis (EDA), we aim to understand how these features relate to diamond prices.
-This analysis will help us answer questions about which features most significantly impact diamond prices and how we can leverage this information to create predictive models.
-Thus, our goal is to develop an accurate model for predicting diamond prices, enabling jewelers to make precise pricing decisions and enhancing the consumer purchasing experience.
+Our dataset contains several essential features for diamond price prediction, such as the aforementioned carat, cut, color, clarity, etc.
+\cite{kaggle} Through exploratory data analysis (EDA), we aim to understand how these features are related to diamond pricing.
+This analysis will help us answer questions about which features most significantly impact diamond prices and ways we can leverage this information to create accurate predictive models.
+Ultimately, our goal is to develop an accurate model for predicting diamond prices, enabling jewelers to make precise pricing decisions and improving the overall consumer purchasing experience.
 This paper details our methodology, including EDA and the application of various machine learning algorithms, to achieve this objective.
 
-The outline of this paper is as follows: Section 2 discusses the historical perspective of machine learning and relevant algorithms as part of the literature review.
+The outline of this paper is as follows: Section 2 reviews the related literature on diamond price prediction and provides a historical perspective on machine learning and relevant algorithms.
 Section 3 presents data description and exploratory data analysis.
-Sections 4 and 5 discuss the proposed methodology and results.
+Sections 4 and 5 highlight the proposed methodology and results.
 Finally, the paper concludes with a discussion and future perspective in Section 6.
@@
 
-Machine learning is a rapidly developing field with a diverse range of algorithms tailored for various tasks.
-This section details important concepts surrounding the ML algorithms we will use.
-In particular, we discuss the historical context \& purpose of regression models, decision trees, and neural networks.
+Much literature has been published to develop predictive models for accurately estimating diamond prices, using various social contexts and individual diamond characteristics.
+Alsuraihi et al.
+\cite{Alsuraihi} aimed to develop an accurate algorithm to estimate diamond prices.
+They considered various features such as diamond sizes and other key factors.
+Various machine learning methods were tested, including Linear Regression, Random Forest Regression, Polynomial Regression, Gradient Descent, and Neural Networks.
+Similarly, Mamonov and Triantoro \cite{Mamonov} studied the relationship between a diamond's physical attributes and respective prices in e-commerce contexts, with the goal of understanding how these attributes affect diamond prices.
+However, their study did not account for the diamond cut, a significant factor affecting market value, which is included in our dataset.
+Pandey et al.
+\cite{Pandey} addressed the challenge of forecasting future values of precious metals like gold and diamonds.
+They used ensemble approaches combined with feature selection techniques to enhance prediction accuracy.
+In an alternate perspective, Scott and Yelowitz \cite{Scott} investigated diamond prices in the context its social status and intrinsic value.
+They collected data from online diamond sellers and empirically examined factors influencing diamond prices, considering carat weight, color, cut, and clarity in determining the logarithm of price.
+Clearly, a variety of approaches have been taken to create accurate prediction models, ranging from investigating social contexts to individual characteristics.
+In this paper, we utilize several algorithms to create accurate models ourselves.
+The latter half of this section provides an overview of key machine learning algorithms of this paper, including regression models, decision trees, and artificial neural networks.
+We examine the Grid Search method for hyperparameter tuning.
 
@@
 
-The most basic prediction model we explore is linear regression.
-This model fits a straight line between the response variable and its predictors, making it extremely simple and interpretable in terms of estimating the regression coefficients \cite{b10}.
-One major benefit of linear regression, compared to more complex models, is its efficiency and ease of interpretation, especially when the relationship between the dependent variable $Y$ and the independent variable(s) $X$ is approximately linear.
-For a single predictor $X$, the general form of a linear regression model is:
-
-\[
-    Y = \beta_0 + \beta_1 X + \epsilon
-\]
-
-Here, $\beta_0$ represents the intercept term, $\beta_1$ is the coefficient for X, and $\epsilon_1 \sim \mathcal{N}(0,\sigma_1^2)$.
-Note that $\epsilon_1$ is independent of the other predictor terms.
-Linear regression relies on several key assumptions, including linearity, independence, homoscedasticity, and normality of residuals \cite{b8}.
-The performance of a linear regression model can be evaluated using metrics such as $R^2$ and Mean Squared Error (MSE).
-Linear regression can be extended to more complex models, such as multiple linear regression, polynomial regression, and regularization methods (e.g., Ridge and Lasso regression) to handle more complex relationships and preventing overfitting.
-
-A crucial part of linear regression is the minimization of the Residual Sum of Squares (RSS). The RSS measures the discrepancy between the observed data and the values predicted by the model:
+Linear regression is one of the simplest and most interpretable prediction models.
+It fits a straight line between the response variable $Y$ and predictors $X$.
+The goal of this model is to minimize the Residual Sum of Squares (RSS) \cite{Seber}:
 
@@
 
-Minimizing the RSS through Ordinary Least Squares (OLS) provides the best-fitting line by ensuring the smallest possible error in predicting the response variable from the predictors.
-For further details on the mathematical foundations of OLS and RSS minimization, refer to Seber and Lee's comprehensive discussion on linear regression \cite{b9}.
-Linear regression has been extensively studied and applied across various fields \cite{b1, b2}.
-Additionally, Kutner et al.
-provide detailed discussions on the statistical assumptions, limitations, and diagnostics of linear regression models \cite{b8}.
+Linear regression assumes linearity, independence, homoscedasticity, and normality of residuals \cite{Kutner}.
+It is evaluated using metrics like $R^2$ and Mean Squared Error (MSE).
+Extensions include multiple linear regression, polynomial regression, and regularization methods like Ridge and Lasso regression to handle complex relationships and prevent overfitting \cite{Wright, Weisberg}.
 
@@
 
-The second major algorithm we review is decision trees.
-Decision trees are powerful tools for classification and regression tasks, especially with tabular data.
-These algorithms split the data into subsets based on the value of input features, creating a tree-like model of decisions.
-Decision trees can be prone to overfitting; however, ensemble methods such as Random Forests (Breiman, 2001) and Gradient Boosting Machines like XGBoost (Chen \& Guestrin, 2016) mitigate this issue by combining multiple trees to improve performance and robustness.
-Random forests, introduced by Breiman (2001), create a "forest" of multiple decision trees and aggregate their predictions.
-This method reduces variance and improves generalization by averaging multiple models .
-XGBoost, a popular implementation of gradient boosting, has been shown to deliver state-of-the-art performance in many machine learning competitions due to its efficiency and scalability.
-
-Despite being recently popularized by the rise of AI applications, neural networks are actually a historical concept developed in the early 1980s.
-It resurgence was sparked by the rise of deep learning, with advancements in computational power to handle larger datasets.
+Decision trees are very effective on tabular data, as it displays strong performance on both regression and classification tasks.
+These models split the data based on feature values, forming a tree-like structure.
+In particular, decision trees are prone to overfitting.
+Ensemble methods, such as Random Forests \cite{Breiman} and Gradient Boosting Machines like XGBoost \cite{Chen}, mitigate this issue by combining multiple trees, thereby enhancing performance and robustness.
 
@@
 
-The foundation of modern neural networks was laid by the introduction of backpropagation by Rumelhart, Hinton, and Williams (1986) .
-Recently, deep learning has revolutionized fields such as image recognition, natural language processing, and game playing.
-LeCun, Bengio, and Hinton (2015) provide a seminal review on the principles and advancements in deep learning.
-A neural network contains many hyperparameters, and there are several elements to finding the best set of such values.
-Grid search is a systematic approach for hyperparameter tuning, where a predefined set of hyperparameters is evaluated exhaustively to identify the optimal configuration for a machine learning model.
-Bergstra and Bengio (2012) introduced grid search as a fundamental method for hyperparameter optimization.
-They highlighted its importance in efficiently exploring the hyperparameter space and provided insights into its limitations, such as scalability and computational cost.
-For a comprehensive discussion on hyperparameter optimization techniques, see the review by Loshchilov and Hutter (2019).
+Neural networks have recently become extremely popular with advances in deep learning and computation ability to handle big data problems \cite{LeCun}.
+These models consist of layers of interconnected nodes (neurons) that learn representations from data.
+Rumelhart et.
+al \cite{Rumelhart} introduced the backpropagation algorithm in order to learn the model weights by minimizing the training loss.
+Modern neural networks use optimizers like Adam (Adaptive Moment Estimation), which combines adaptive learning rates and momentum for efficient training.
 
-The Adam optimizer, short for Adaptive Moment Estimation, is a widely used optimization algorithm for training deep neural networks.
-It combines the advantages of both adaptive learning rate methods and momentum-based optimization techniques, offering fast convergence and robustness to noisy gradients.
-Kingma and Ba (2014) introduced the Adam optimizer, presenting it as an extension of the stochastic gradient descent with adaptive learning rates.
-They demonstrated its effectiveness in training deep learning models across various tasks, including image classification, natural language processing, and reinforcement learning.
-Adam maintains adaptive learning rates for each parameter by computing individual learning rates based on the first and second moments of the gradients.
-This adaptive behavior allows Adam to automatically adjust the learning rate during training, leading to improved convergence and generalization performance.
+\subsection{Grid Search}
 
-In conclusion, grid search and the Adam optimizer are essential components of the machine learning and deep learning toolkit.
-While grid search provides a systematic approach for hyperparameter tuning, Adam offers an efficient optimization algorithm for training deep neural networks.
-Understanding their strengths, limitations, and best practices is crucial for effectively applying these techniques in practical scenarios.
+Both Neural Networks and decision trees include a variety of hyperparameters that significantly impact their performance and overall generalization ability on unseen data.
+Grid search is a method for hyperparameter tuning that evaluates a predefined set (grid) of hyperparameters and finds the optimal configuration among them.
+This method is quite computationally heavy, as it explore every configuration among the given hyperparameter space to identify the most suitable combination for a given problem.
+Bergstra and Bengio (2012) \cite{Bergstra} discussed grid search's efficiency and limitations, highlighting its importance in fine-tuning machine learning models for optimal performance.
+
+Therefore, our literature review provides an overview of the research conducted on predicting diamond prices, including discussions on pertinent algorithms such as linear regression, decision trees, and neural networks.
+For hyperparameter tuning, we plan to utilize Grid Search to optimize the performance of both the decision trees and neural network models.
 
@@
 
-The \emph{diamond prices} dataset contains detailed information about 54,000 diamonds, including their prices and various characteristics \cite{a2}.
+The \emph{diamond prices} dataset contains detailed information about 54,000 diamonds, including their prices and various characteristics \cite{kaggle}.
 The predictor variables in this dataset are as follows: i) \textbf{Carat}: Represents the weight of the diamond, ranging from 0.2 to 5.01 carats.
@@
 vii) \textbf{Table}: Represents the width of the top of the diamond relative to its widest point, ranging from 43\% to 95\%.
-The dataset does not contain any missing values, so we do not need to address missing data in the exploratory data analysis step.
+The dataset does not contain any missing values, so we do not need to any address missing data in the exploratory data analysis step.
 
@@
     \includegraphics[width=0.8\linewidth]{Figure_1.png}
-    \caption{Distribution Plot of Continuous Features. Reveals trends in data such as skewness, location of outliers, etc.}
+    \caption{Distribution Plot of Continuous Features. Identify trends our dataset such as skewness, location of outliers, mean, median, scale, etc.}
     \label{fig:image_label}
@@
 These plots suggest that our dataset may contain a significant number of upper outliers, which we will consider removing in the data pre-processing stage.
-To further gain insight on the distribution of our features, the table below summarizes the mean, median, and standard deviation of the continuous features:
+To further gain insight on the distribution of our continuous features, the table below summarizes the mean, median, and standard deviation of the each predictor:
 
@@
 However, the standard deviation values highlight the variability within each feature, such as carat and dimensions (x, y, z), display considerable spread.
-This variability and potential skewness, especially in the carat feature from the plots, could influence our predictive algorithms.
+This variability and potential skewness (from the plots), especially in the carat feature from the plots, could influence our predictive algorithms.
 Next, we will create pairwise scatter-plots to identify trends between diamond prices and the features.
@@
 
-In this paper, we explore the predictive capabilities of several machine learning algorithms, ranging from linear regression and polynomial regression to decision trees and artificial neural networks.
-Our approach involves multiple steps to build and evaluate models that can accurately predict diamond prices based on various features.
+In this paper, we will explore the predictive capabilities of several machine learning algorithms, ranging from linear regression, decision trees, and artificial neural networks.
+Our approach involved several steps to develop and evaluate model performance to choose the most accurate algorithm.
+Data preprocessing has already been completed as described above.
 
-We begin with traditional linear regression to fit a straight-line model between diamond prices and the aforementioned features.
+\subsection{Linear Regression}
+
+We begin with linear regression model to fit a straight-line model between diamond prices and the aforementioned features.
 This initial step helps us gain intuition about the relationship between the features and diamond prices using a simple model.
-Recognizing that the price exhibited a quadratic relationship with carat and diamond dimensions, we extend our analysis to polynomial regression by including squared terms of these features.
-This allows us to capture potential nonlinear relationships and assess whether polynomial regression outperforms traditional linear regression.
+From our EDA step, since price displayed a quadratic relationship with carat and diamond dimensions (x,y,z), we also test the performance of a polynomial regression model by including squared terms of these respective features.
+Thus, we can capture potential nonlinear relationships and assess whether polynomial regression outperforms traditional linear regression.
 
-For decision tree-based models, we employ two powerful algorithms: Random Forests and XGBoost.
-Random Forests aggregate the predictions of multiple decision trees to enhance accuracy and reduce overfitting.
-XGBoost, a gradient boosting algorithm, further refines predictions by optimizing the model iteratively to minimize errors.
-These models are well-suited for capturing complex interactions between features and are expected to perform robustly on our dataset.
+\subsection{Decision Trees}
 
-Finally, we explore the use of artificial neural networks (ANNs) for generating predictions.
-To optimize our ANN, we conduct a grid search for hyperparameter tuning, ensuring we select the best set of parameters for our model.
-Our neural network architecture consists of a 3-layer ANN with 64 neurons in the first hidden layer, using the ReLU activation function.
-The output layer also employs the ReLU activation function, as our task involves regression.
+For decision tree-based models, we employ two algorithms; Random Forests and XGBoost.
+Random Forests aggregate the predictions of multiple decision trees to enhance accuracy and reduce overfitting; XGBoost is a gradient boosting algorithm that iteratively optimizes the model to minimize overall error.
+We implement grid search with a 5-fold cross-validation to identify the optimal set of hyperparameters.
+These are the following set of hyper-parameters we tested:
 
-To evaluate the performance of our models, we will compare their results using several metrics, including Mean Squared Error (MSE), Mean Absolute Error (MAE), and R-squared (R²).
-These metrics will help us determine the accuracy and reliability of each model on testing data.
-By comparing these measures, we aim to identify the most effective algorithm for predicting diamond prices, providing valuable insights for jewelers and marketers in the industry.
+\begin{itemize}
+    \item \textbf{Random Forests}: Our parameter grid explores options for 'bootstrap' (set to True), 'max\_features' ('auto' and 'sqrt'), 'max\_depth' (None, 10, and 20), 'min\_samples\_leaf' (1 and 2), 'min\_samples\_split' (2 and 5), and 'n\_estimators' (100, 200, and 300).
+    \item \textbf{XGBoost}: We examine hyperparameters such as 'max\_depth' (3, 6, and 9), 'learning\_rate' (0.1 and 0.01), 'n\_estimators' (100, 200, and 300), 'subsample' (0.8 and 1.0), 'colsample\_bytree' (0.8 and 1.0), 'gamma' (0 and 0.1), 'reg\_alpha' (0 and 0.1), and 'reg\_lambda' (0 and 0.1).
+\end{itemize}
 
-Through this comprehensive methodology, we seek to develop an accurate and robust model for predicting diamond prices, leveraging the strengths of various machine learning techniques to achieve the best possible performance.
+This broad range of values for each hyperparameter enables us to explore a wide array of model configurations and identify the optimal combination for our dataset.
+Subsequently, we train the model using this set of hyperparameters on the training data and assess its performance on the testing set.
+The plot of training and testing loss vs.\ epoch is provided to demonstrate model performance.
+
+\subsection{Artificial Neural Networks}
+
+In our exploration of predicting diamond prices, we employ artificial neural networks (ANNs).
+Our ANN architecture consists of a single hidden layer with ReLU activation, followed by an output layer also utilizing ReLU activation, as our task involves regression.
+To optimize the performance of our ANN, we conduct a grid search for hyperparameter tuning, aiming to select the most suitable parameters for our model.
+We explore the following hyperparameters during grid search with a 5-fold cross-validation procedure.
+We explore the following set of hyperparameters:
+
+\begin{itemize}
+    \item 'learning\_rate' (0.001, 0.0001)
+    \item 'momentum' (0.5, 0.9)
+    \item 'number of neurons in hidden layer' (16, 32, 64)
+    \item 'dropout\_rate' (None, 0.2, 0.3)
+    \item 'batch\_size' (256, 512, 1028)
+\end{itemize}
+
+By evaluating these hyperparameters, we will to identify the optimal configuration of hyperparameters that maximizes the predictive performance of our ANN model on the given dataset.
+Additionally, we incorporate the Adam optimizer into our ANN model to further improve the training process and overall predictive accuracy.
+
+\subsection{Evaluating Model Performance}
+
+In our grid search, we will use the \emph{negative mean squared error} as our metric to identify the optimal set of hyperparameters for both decision trees and the artificial neural network.
+To assess the performance of our models on the testing data, we will compare their results using several metrics, including Mean Squared Error (MSE), Mean Absolute Error (MAE), and R-squared ($R^2$) between the true values and our predictions.
+These metrics will help evaluating the accuracy and fit of each algorithm on the testing data.
+By analyzing these measures, our goal is to find the most effective algorithm for predicting diamond prices as it offers valuable insights for jewelers and marketers in the industry.
+Through this methodology, we will develop an accurate for predicting diamond prices, leveraging the strengths of various machine learning techniques and achieve best performance possible.
 
 \section{Experimental results and evaluation}
+
+Our experimental methodology provides nuanced results across each of the machine learning algorithms. In particular, the results are as follows:
+
+\subsection{Linear \& Polynomial Regression}
+
+In terms of our training data, we tested the effectivness of both linear and polynomial regression models on our training data using 5-Fold Cross validation.
+Based on the results, this was our mean score error on training data:
+
+\begin{table}[h]
+\centering
+\begin{tabular}{|c|c|c|}
+\hline
+Column 1 & Column 2 & Column 3 \\
+\hline
+Row 1 & Cell 1 & Cell 2 \\
+Row 2 & Cell 3 & Cell 4 \\
+Row 3 & Cell 5 & Cell 6 \\
+Row 4 & Cell 7 & Cell 8 \\
+\hline
+\end{tabular}
+\caption{A simple table with 3 columns and 4 rows.}
+\label{tab:mytable}
+\end{table}
+
+\subsection{Decision Trees}
+
+\textbf{Random Forest Algorthm}
+
+For our decision tree algorirhtms, the results of grid searhc provided the following best set of hyperpatemrs for our random forest algorhtm.
+
+\begin{table}[h]
+\centering
+\begin{tabular}{|c|c|c|}
+\hline
+Column 1 & Column 2 & Column 3 \\
+\hline
+Row 1 & Cell 1 & Cell 2 \\
+Row 2 & Cell 3 & Cell 4 \\
+Row 3 & Cell 5 & Cell 6 \\
+Row 4 & Cell 7 & Cell 8 \\
+\hline
+\end{tabular}
+\caption{A simple table with 3 columns and 4 rows.}
+\label{tab:mytable}
+\end{table}
+
+\textbf{XGBoost}
+
+\begin{table}[h]
+\centering
+\begin{tabular}{|c|c|c|}
+\hline
+Column 1 & Column 2 & Column 3 \\
+\hline
+Row 1 & Cell 1 & Cell 2 \\
+Row 2 & Cell 3 & Cell 4 \\
+Row 3 & Cell 5 & Cell 6 \\
+Row 4 & Cell 7 & Cell 8 \\
+\hline
+\end{tabular}
+\caption{A simple table with 3 columns and 4 rows.}
+\label{tab:mytable}
+\end{table}
+
+\subsection{Neural Network}
+
+The grid search for neural network presents the following results:
+
+\begin{table}[h]
+\centering
+\begin{tabular}{|c|c|c|}
+\hline
+Column 1 & Column 2 & Column 3 \\
+\hline
+Row 1 & Cell 1 & Cell 2 \\
+Row 2 & Cell 3 & Cell 4 \\
+Row 3 & Cell 5 & Cell 6 \\
+Row 4 & Cell 7 & Cell 8 \\
+\hline
+\end{tabular}
+\caption{A simple table with 3 columns and 4 rows.}
+\label{tab:mytable}
+\end{table}
+
+\subsection{Comparison of models on testing data}
 
@@
 \begin{thebibliography}{00}
-\bibitem{a1} Kino, S., Hsu, Y.-T., Shiba, K., Chien, Y.-S., Mita, C., Kawachi, I., and Daoud, A.
+\bibitem{kino} Kino, S., Hsu, Y.-T., Shiba, K., Chien, Y.-S., Mita, C., Kawachi, I., and Daoud, A.
 (2021).
@@
 SSM - Population Health, 15, 100836.
-\bibitem{c1} Sarker, I.H. Machine Learning: Algorithms, Real-World Applications and Research Directions. SN COMPUT. SCI. 2, 160 (2021). https://doi.org/10.1007/s42979-021-00592-x
-\bibitem{a2} Agrawal, S. (2017a, May 25). Diamonds. Kaggle. https://www.kaggle.com/datasets/shivam2503/diamonds 
-\bibitem{a3} U.S.
+\bibitem{sarker} Sarker, I.H. Machine Learning: Algorithms, Real-World Applications and Research Directions. SN COMPUT. SCI. 2, 160 (2021). https://doi.org/10.1007/s42979-021-00592-x
+\bibitem{Alsuraihi} Alsuraihi, W., Al-hazmi, E., Bawazeer, K., \& AlGhamdi, H.
+Machine learning algorithms for diamond price prediction.
+In Proceedings of the 2020 2nd International Conference on Image, Video and Signal Processing, 150–154 (2020).
+\bibitem{Mamonov} Mamonov, S. \& Triantoro, T. Subjectivity of diamond prices in online retail: Insights from a data mining study. J. Theor. Appl. Electron. Commer. Res. 13(2), 15–28 (2018).
+\bibitem{Pandey} Pandey, A.
+C., Misra, S., \& Saxena, M.
+Gold and diamond price prediction using enhanced ensemble learning.
+In 2019 Twelfth International Conference on Contemporary Computing (IC3), 1–4 (IEEE, 2019).
+\bibitem{Scott} Scott, F. \& Yelowitz, A. Pricing anomalies in the market for diamonds: Evidence of conformist behavior. Econ. Inq. 48(2), 353–368 (2010).
+\bibitem{Kutner}Michael Kutner, Christopher Nachtsheim, John Neter, and William Li. Applied Linear Statistical Models. 1974.
+\bibitem{Seber}Seber, G. A. F., \& Lee, A. J. (2012). Linear Regression Analysis. Hoboken, NJ: John Wiley \& Sons.
+\bibitem{Wright} Wright, John T. "Linear Regression Analysis." British Medical Journal, vol. 310, no. 6977, 1995, pp. 1120-1124. BMJ Group
+\bibitem{Weisberg} Weisberg, S. (2005). Applied Linear Regression. Wiley.
+\bibitem{Breiman} Breiman, L. (2001). Random forests. Machine Learning, 45(1), 5-32.
+\bibitem{Chen} Chen, T., \& Guestrin, C.
+(2016).
+XGBoost: A Scalable Tree Boosting System.
+In Proceedings of the 22nd ACM SIGKDD International Conference on Knowledge Discovery and Data Mining (pp.
+785-794).
+\bibitem{LeCun} LeCun, Y., Bengio, Y., \& Hinton, G. (2015). Deep learning. Nature, 521(7553), 436-444.
+\bibitem{Rumelhart} Rumelhart, D. E., Hinton, G. E., \& Williams, R. J. (1986). Learning representations by back-propagating errors. Nature, 323(6088), 533-536.
+\bibitem{Bergstra} Bergstra, J., \& Bengio, Y. (2012). Random search for hyper-parameter optimization. Journal of Machine Learning Research, 13(Feb), 281-305.
+\bibitem{us} U.S.
 jewelry market size and share: Industry Report, 2030.
@@
 https://www.grandviewresearch.com/industry-analysis/us-jewelry-market-report
-\bibitem{a4} Diamond Jewelry Market Size \& Share Analysis Report, 2030. (n.d.). https://www.grandviewresearch.com/industry-analysis/diamond-jewelry-market-report 
-\bibitem{b1} Weisberg, S. (2005). Applied Linear Regression. Wiley.
-\bibitem{b2} Chatterjee, S., \& Hadi, A. S. (2015). Regression Analysis by Example. Wiley.
-\bibitem{b3} Breiman, L. (2001). Random Forests. Machine Learning, 45(1), 5-32.
-\bibitem{b4} Chen, T., \& Guestrin, C.
-(2016).
-XGBoost: A Scalable Tree Boosting System.
-In Proceedings of the 22nd ACM SIGKDD International Conference on Knowledge Discovery and Data Mining (pp.
-785-794).
-\bibitem{b5} Rumelhart, D. E., Hinton, G. E., \& Williams, R. J. (1986). Learning representations by back-propagating errors. Nature, 323(6088), 533-536.
-\bibitem{b6} LeCun, Y., Bengio, Y., \& Hinton, G. (2015). Deep learning. Nature, 521(7553), 436-444.
-\bibitem{b7}Vaswani, A., Shazeer, N., Parmar, N., Uszkoreit, J., Jones, L., Gomez, A.
-N., Kaiser, Ł., \& Polosukhin, I.
-(2017).
-Attention is all you need.
-In Advances in Neural Information Processing Systems (pp.
-5998-6008).
-\bibitem{b8}Michael Kutner, Christopher Nachtsheim, John Neter, and William Li. Applied Linear Statistical Models. 1974.
-\bibitem{b9}Seber, G. A. F., \& Lee, A. J. (2012). Linear Regression Analysis. Hoboken, NJ: John Wiley \& Sons. 
-\bibitem{b10} Wright, John T. "Linear Regression Analysis." British Medical Journal, vol. 310, no. 6977, 1995, pp. 1120-1124. BMJ Group
+\bibitem{diamond} Diamond Jewelry Market Size \& Share Analysis Report, 2030. (n.d.). https://www.grandviewresearch.com/industry-analysis/diamond-jewelry-market-report
+\bibitem{kaggle} Agrawal, S. (2017a, May 25). Diamonds. Kaggle. https://www.kaggle.com/datasets/shivam2503/diamonds
 \end{thebibliography}

### 2024-06-03 code/exploratory_data_analysis.py
exploratory data analysis & data pre-processing

@@ -1,0 +1,152 @@
+# import necessary modules
+import pandas as pd
+import numpy as np
+import matplotlib.pyplot as plt
+import seaborn as sns
+
+from sklearn.linear_model import LinearRegression
+from sklearn.preprocessing import PolynomialFeatures
+from sklearn.preprocessing import StandardScaler
+from sklearn.model_selection import train_test_split, GridSearchCV, KFold, cross_val_score
+from sklearn.metrics import mean_squared_error, r2_score
+from sklearn.ensemble import RandomForestRegressor
+import xgboost as xgb
+
+import tensorflow as tf
+from tensorflow.keras.models import Sequential
+from tensorflow.keras.layers import Dense, Dropout
+from tensorflow.keras import regularizers
+from tensorflow.keras.optimizers.legacy import Adam
+from tensorflow.keras.wrappers.scikit_learn import KerasRegressor
+
+### Exploratory Data Analysis
+### ----------------------------------------------------------------------------
+df = pd.read_csv("~/Desktop/Year_three/Spring_2024/ECS171/ECS171_Final/diamonds.csv") 
+df = df.iloc[:, 1:]  # remove index column          
+# df.head()
+
+### Plot distribution of continous features
+features_to_plot = ['carat', 'depth', 'table', 'x','y','z']
+plt.figure(figsize=(8, 6))
+for i, feature in enumerate(features_to_plot, 1):
+    plt.subplot(2, 3, i)
+    sns.histplot(df[feature], kde=True)
+    plt.title(feature)
+
+plt.tight_layout()
+plt.show()
+print("\nmean:\n{} , \nmedian:\n{} , \nstd:\n {}".format(df[features_to_plot].mean() , 
+                                                         df[features_to_plot].median(), 
+                                                         df[features_to_plot].std()))
+
+### Plot scatterplot of continous features vs. price
+feature_names = ['carat', 'depth', 'table', 'x','y','z']
+y = df['price']
+plt.figure(figsize=(15, 10))
+for i, feature in enumerate(feature_names):
+    plt.subplot(3, 2, i+1)
[…]

### 2024-06-03 code/regression_models.py
create regression models

@@ -1,0 +1,55 @@
+### Assuming we have X, X_test, y, y_test from exploratory_data_analysis.py
+
+### Linear Regression
+### ----------------------------------------------------------------------------
+
+### Cross validation procedure
+model = LinearRegression()
+kf = KFold(n_splits=5, shuffle=True, random_state=42) 
+
+mse_scores = cross_val_score(model, X, y, cv=kf, scoring='neg_mean_squared_error')
+mse_scores = -mse_scores
+mean_mse = np.mean(mse_scores)
+std_mse = np.std(mse_scores)
+print(f'Mean MSE: {mean_mse}')
+print(f'Standard Deviation of MSE: {std_mse}')
+
+
+### Model accuracy on testing data
+model = LinearRegression()
+model.fit(X, y)
+y_pred = model.predict(X_test)
+mae = np.mean(abs(y_pred - y_test))
+r_squared = r2_score(y_test, y_pred)
+mse = mean_squared_error(y_test, y_pred)
+print(f'MAE: {mae}')
+print("R-squared:", r_squared)
+print("Mean Squared Error:", mse)
+
+
+### Polynomial Regression
+### ----------------------------------------------------------------------------
+poly = PolynomialFeatures(2)
+X_poly = poly.fit_transform(X)
+X_poly_test = poly.fit_transform(X_test)
+
+# Cross validation procedure
+model = LinearRegression()
+kf = KFold(n_splits=10, shuffle=True, random_state=42) # 5 fold CV
+mse_scores = cross_val_score(model, X_poly, y, cv=kf, scoring='neg_mean_squared_error')
+mse_scores = -mse_scores
+mean_mse = np.mean(mse_scores)
+std_mse = np.std(mse_scores)
+print(f'Mean MSE: {mean_mse}')
+print(f'Standard Deviation of MSE: {std_mse}')
+
+# model fit on testing data
+model = LinearRegression()
+model.fit(X_poly, y)
+y_pred = model.predict(X_poly_test)
+mae = np.mean(abs(y_pred - y_test))
+r_squared = r2_score(y_test, y_pred)
+mse = mean_squared_error(y_test, y_pred)
+print(f'MAE: {mae}')
+print("R-squared:", r_squared)
+print("Mean Squared Error:", mse)

### 2024-06-03 code/decision_trees.py
Create decision_trees.py

@@ -1,0 +1,87 @@
+### Assuming we have X, X_test, y, y_test from the exploratory_data_analysis.py
+
+### Random Forests
+### ----------------------------------------------------------------------------
+### grid search
+rf = RandomForestRegressor()  
+
+rf_param_grid = {
+    'bootstrap': [True],
+    'max_features': ['auto', 'sqrt'],
+    'max_depth': [None, 10, 20],
+    'min_samples_leaf': [1, 2],
+    'min_samples_split': [2, 5],
+    'n_estimators': [100, 200, 300]
+}
+
+grid_search = GridSearchCV(estimator=rf, param_grid=rf_param_grid, cv=5, n_jobs=-1, verbose=0, scoring='neg_mean_squared_error')
+grid_search.fit(X, y)  
+
+print("Best Parameters:", grid_search.best_params_)
+print("Best Score:", grid_search.best_score_)
+
+### model fit on testing data
+best_params = {
+    'bootstrap': True,
+    'max_depth': 20,
+    'max_features': 'sqrt',
+    'min_samples_leaf': 1,
+    'min_samples_split': 2,
+    'n_estimators': 300
+}
+
+rf_regressor_best = RandomForestRegressor(**best_params, random_state=42)
+rf_regressor_best.fit(X, y)
+
+y_pred = rf_regressor_best.predict(X_test)
+mae = np.mean(abs(y_pred - y_test))
+r_squared = r2_score(y_test, y_pred)
+mse = mean_squared_error(y_test, y_pred)
+print(f'MAE: {mae}')
+print("R-squared:", r_squared)
+print("Mean Squared Error:", mse)
+
+### XGBoost
+### ----------------------------------------------------------------------------
+### grid search
+
+xgb_model = xgb.XGBRegressor()  
+xgb_param_grid = {
+    'max_depth': [3, 6, 9],  
+    'learning_rate': [0.1, 0.01],  
+    'n_estimators': [100, 200, 300],  
+    'subsample': [0.8, 1.0],  
+    'colsample_bytree': [0.8, 1.0],  
+    'gamma': [0, 0.1],  
+    'reg_alpha': [0, 0.1],  
+    'reg_lambda': [0, 0.1]  
+}
+grid_search_xgb = GridSearchCV(estimator=xgb_model, param_grid=xgb_param_grid, cv=5, scoring='neg_mean_squared_error', verbose=0)
+grid_search_xgb.fit(X, y)  
[…]

### 2024-06-03 code/neural_network.py
Create neural_network.py

@@ -1,0 +1,56 @@
+### Assuming we have X, X_test, y, y_test from exploratory_data_analysis.py
+
+### Defining the Neural Network ------------------------------------------------
+### Grid Search to find best hyper-params
+def create_model(num_layers = 3, num_neurons = 64, dropout_rate=0.0, optimizer='adam', learning_rate = 0.01):
+    model = Sequential()
+
+    # hidden layer one
+    model.add(Dense(num_neurons, input_dim=9, activation='relu'))  
+    model.add(Dropout(dropout_rate))
+    
+    # following hidden layers
+    for _ in range(num_layers - 1):
+        model.add(Dense(num_neurons, activation='relu'))
+        model.add(Dropout(dropout_rate))
+    
+    # output layer
+    model.add(Dense(1, activation='relu'))
+    
+    # optimizer
+    if optimizer == 'rmsprop':
+        opt = tf.keras.optimizers.legacy.RMSprop(learning_rate=learning_rate)
+    elif optimizer == 'adam':
+        opt = tf.keras.optimizers.legacy.Adam(learning_rate=learning_rate)
+    elif optimizer == 'SGD':
+        opt = tf.keras.optimizers.legacy.SGD(learning_rate=learning_rate)
+    else:
+        raise ValueError('optimizer {} unrecognized'.format(optimizer))
+        
+    # compile
+    network.compile(optimizer = opt, 
+                loss = 'mean_squared_error', 
+                metrics=['mae', tf.keras.metrics.RootMeanSquaredError()])
+    
+    return model
+
+### Implement Grid Search -------------------------------------------------------
+ann = KerasRegressor(build_fn=create_model, verbose=0)
+
+# these are the values for the other hyperparameters
+param_grid = dict(
+    num_layers= [1, 3, 5, 8],
+    num_neurons= [16, 32, 64],
+    dropout_rate= [0.2, 0.5],
+    optimizer = ['rmsprop', 'adam', 'SGD'],
+    batch_size = [32, 64, 128, 256, 512, 1024, 2048],
+    epochs = [10, 25, 50, 100, 150, 200],
+    learning_rate = [0.001, 0.01, 0.1, 0.2, 0.3]
+)
+
+# apply 5-fold CV to grid search
+grid = GridSearchCV(estimator=ann, cv=5, param_grid=param_grid)
+grid_result = grid.fit(X, y)
+
+# print best set of parameters // comes after the warnings 
+print("Best: %f using %s" % (grid_result.best_score_, grid_result.best_params_))

### 2024-06-03 paper/main.tex
Update main.tex

@@
 This analysis will help us answer questions about which features most significantly impact diamond prices and ways we can leverage this information to create accurate predictive models.
-Ultimately, our goal is to develop an accurate model for predicting diamond prices, enabling jewelers to make precise pricing decisions and improving the overall consumer purchasing experience.
+Ultimately, our goal is to develop an accurate model for predicting diamond prices, enabling jewelers to make precise pricing decisions and improve the overall consumer purchasing experience.
 This paper details our methodology, including EDA and the application of various machine learning algorithms, to achieve this objective.
@@
 Clearly, a variety of approaches have been taken to create accurate prediction models, ranging from investigating social contexts to individual characteristics.
-In this paper, we utilize several algorithms to create accurate models ourselves.
-The latter half of this section provides an overview of key machine learning algorithms of this paper, including regression models, decision trees, and artificial neural networks.
-We examine the Grid Search method for hyperparameter tuning.
+In this paper, we utilize several algorithms to create accurate models by building on the previous work.
+With this, the latter half of this section provides an overview of several machine learning algorithms applied in this paper, including regression models, decision trees, and neural networks.
+We also examine Grid Search as a method for hyperparameter tuning.
 
@@
 
-Linear regression is one of the simplest and most interpretable prediction models.
-It fits a straight line between the response variable $Y$ and predictors $X$.
+Linear regression is one of the simplest and one of the most interpretable prediction models.
+It fits a straight line between the response variable $Y$ and its predictors $X$.
 The goal of this model is to minimize the Residual Sum of Squares (RSS) \cite{Seber}:
@@
 
-Linear regression assumes linearity, independence, homoscedasticity, and normality of residuals \cite{Kutner}.
-It is evaluated using metrics like $R^2$ and Mean Squared Error (MSE).
-Extensions include multiple linear regression, polynomial regression, and regularization methods like Ridge and Lasso regression to handle complex relationships and prevent overfitting \cite{Wright, Weisberg}.
+Linear regression requires several assumptions: linearity, independence, homoscedasticity, and normality of residuals \cite{Kutner}.
+Its performance may be  evaluated using metrics like $R^2$, Mean Squared Error (MSE), Mean Absolute Prediction Error (MAE).
+Extensions include multiple linear regression, polynomial regression, and regularization methods like Ridge and Lasso regression to handle complex relationships and handle overfitting \cite{Wright, Weisberg}.
 
@@
 
-Decision trees are very effective on tabular data, as it displays strong performance on both regression and classification tasks.
+Decision trees are very effective on tabular data, as they display strong performance on both regression and classification tasks.
 These models split the data based on feature values, forming a tree-like structure.
 In particular, decision trees are prone to overfitting.
-Ensemble methods, such as Random Forests \cite{Breiman} and Gradient Boosting Machines like XGBoost \cite{Chen}, mitigate this issue by combining multiple trees, thereby enhancing performance and robustness.
+Ensemble methods, such as Random Forests \cite{Breiman} and Gradient Boosting Machines like XGBoost \cite{Chen}, mitigate this issue by combining multiple trees, thereby enhancing performance and robustness of these methods.
 
@@
 These models consist of layers of interconnected nodes (neurons) that learn representations from data.
-Rumelhart et.
-al \cite{Rumelhart} introduced the backpropagation algorithm in order to learn the model weights by minimizing the training loss.
-Modern neural networks use optimizers like Adam (Adaptive Moment Estimation), which combines adaptive learning rates and momentum for efficient training.
+In forward propagation, input data traverses through neural network layers, where each layer applies weights and passes results through activation functions, crucial for introducing nonlinearity.
+Activation functions like ReLU facilitate this nonlinearity, enabling neural networks to learn complex patterns and relationships in data, enhancing their effectiveness in various tasks.
+To learn the model weights, Rumelhart et.
+al \cite{Rumelhart} introduced the backpropagation algorithm by iteratively minimizing the training loss across each epoch.
+SGD (Stochastic Gradient Descent) is a foundational optimizer that updates parameters in the direction of the negative gradient of the loss function, often with momentum to accelerate convergence \cite{Bottou}.
+Modern neural networks use optimizers like Adam (Adaptive Moment Estimation), which combines adaptive learning rates and momentum for efficient training \cite{Kingma}.
+RMSprop (Root Mean Square Propagation) is another popular optimizer in modern neural networks, which adaptively scales the learning rate for each parameter based on the magnitude of recent gradients \cite{Tieleman}.
+We test each of these optimizers when finding the optimal set of hyperparameters of our neural network model.
 
@@
 Both Neural Networks and decision trees include a variety of hyperparameters that significantly impact their performance and overall generalization ability on unseen data.
+In the context of decision trees, important hyperparameters include the maximum depth of the tree, minimum samples required to split a node, and the minimum samples required at each leaf node.
+For neural networks, key hyperparameters encompass the learning rate, batch size, number of hidden layers, and the number of neurons per layer.
 Grid search is a method for hyperparameter tuning that evaluates a predefined set (grid) of hyperparameters and finds the optimal configuration among them.
 This method is quite computationally heavy, as it explore every configuration among the given hyperparameter space to identify the most suitable combination for a given problem.
-Bergstra and Bengio (2012) \cite{Bergstra} discussed grid search's efficiency and limitations, highlighting its importance in fine-tuning machine learning models for optimal performance.
+Bergstra and Bengio (2012) \cite{Bergstra} discusses grid search's efficiency and limitations highlights its importance in fine-tuning machine learning models for optimal performance.
 
-Therefore, our literature review provides an overview of the research conducted on predicting diamond prices, including discussions on pertinent algorithms such as linear regression, decision trees, and neural networks.
+With this, our literature review provides an overview of the research conducted on predicting diamond prices, including discussions on pertinent algorithms such as linear regression, decision trees, and neural networks.
 For hyperparameter tuning, we plan to utilize Grid Search to optimize the performance of both the decision trees and neural network models.
+Next, we proceed with dataset exploration and perform pre-processing on our features.
 
@@
     \centering
-    \includegraphics[width=0.8\linewidth]{Figure_1.png}
-    \caption{Distribution Plot of Continuous Features. Identify trends our dataset such as skewness, location of outliers, mean, median, scale, etc.}
+    \includegraphics[width=0.745\linewidth]{Figure_1.png}
+    \caption{Distribution Plot of Continuous Features. Identify trends in our dataset such as skewness, location of outliers, mean, median, scale, etc.}
     \label{fig:image_label}
@@
 
-From the scatterplots of each feature against the response variable, we observe a quadratic relationship between the features, carat, x, y, and z, with respect to the the response variable price.
+From the scatter-plots of each feature against the response variable, we observe a quadratic relationship between the features, carat, x, y, and z, with respect to the the response variable price.
 This suggests that a polynomial regression model might be suitable for our dataset in addition to a linear regression model.
-Thus, we will evaluate the polynomial regression model's performance on the dataset compared to traditional linear regression.
-Next, we will also analyze the correlations between these continuous features and the target variable, price, to determine the most influential predictors.
-This will allow us to identify the most relevant features and also potential multicollinearity and interactions.
+Thus, we will evaluate the polynomial regression model's performance on this dataset and compare it with traditional linear regression.
+Next, we will analyze the correlations between the continuous features and the target variable, price, to determine the most influential predictors.
+This will allow us to identify the most relevant features and potential multicollinearity and interactions among the predictors.
 
@@
     \includegraphics[width=0.8\linewidth]{Figure_3.png} % Replace "example-image" with the filename of your image
-    \caption{Correlation Heat-Map for Continous Features vs Diamond Price}
+    \caption{Correlation Heat-Map for Continuous Features vs Diamond Price}
     \label{fig:image_label}
@@
 Thus, this plot suggests that Carat (and potentially dimensions) are the strongest predictors for diamond price.
-We showcase data preprocessing methods for outlier removal using carat below.
 On the other hand, both depth percentage and table have almost zero correlation, indicating that they do not have much association with diamond prices.
+We showcase data pre-processing methods for outlier removal using carat below.
 Thus, this heatmap is insightful in identifying the key predictors for diamond prices and guiding our feature selection process.
@@
     \includegraphics[width=0.8\linewidth]{figure_5.png} % Replace "example-image" with the filename of your image
-    \caption{Historgram for the distribution of categorical features: Cut, Clarity, and Color }
+    \caption{Histogram for the distribution of categorical features: Cut, Clarity, and Color }
     \label{fig:image_label}
@@
 Our approach involved several steps to develop and evaluate model performance to choose the most accurate algorithm.
-Data preprocessing has already been completed as described above.
+Data pre-processing has already been completed as described above.
+We divided the dataset into training and testing sets using a 75:25 split.
+The training data consists of 38,000 rows, while the testing data contains 12,000 rows.
+We use cross validation procedures (detailed below) to find the optimal degree of our polynomial regression model and set of optimal hyperparameters for the decision trees \& neural network models.
 
@@
 This initial step helps us gain intuition about the relationship between the features and diamond prices using a simple model.
-From our EDA step, since price displayed a quadratic relationship with carat and diamond dimensions (x,y,z), we also test the performance of a polynomial regression model by including squared terms of these respective features.
-Thus, we can capture potential nonlinear relationships and assess whether polynomial regression outperforms traditional linear regression.
+From our EDA step, since price displayed a quadratic relationship with carat and diamond dimensions (x,y,z), we implement a polynomial regression model to test its performance.
+While the graph indicates a potential quadratic relationship, we employ 5-fold cross-validation by testing powers from 1 to find the optimal degree of our regression model.This allows us to capture nonlinear relationships and assess whether polynomial regression outperforms traditional linear regression.
 
@@
 
-For decision tree-based models, we employ two algorithms; Random Forests and XGBoost.
+For decision tree-based models, we employ two algorithms: Random Forests and XGBoost.
 Random Forests aggregate the predictions of multiple decision trees to enhance accuracy and reduce overfitting; XGBoost is a gradient boosting algorithm that iteratively optimizes the model to minimize overall error.
 We implement grid search with a 5-fold cross-validation to identify the optimal set of hyperparameters.
-These are the following set of hyper-parameters we tested:
+These are the following set of hyperparameters we tested:
 
 \begin{itemize}
-    \item \textbf{Random Forests}: Our parameter grid explores options for 'bootstrap' (set to True), 'max\_features' ('auto' and 'sqrt'), 'max\_depth' (None, 10, and 20), 'min\_samples\_leaf' (1 and 2), 'min\_samples\_split' (2 and 5), and 'n\_estimators' (100, 200, and 300).
-    \item \textbf{XGBoost}: We examine hyperparameters such as 'max\_depth' (3, 6, and 9), 'learning\_rate' (0.1 and 0.01), 'n\_estimators' (100, 200, and 300), 'subsample' (0.8 and 1.0), 'colsample\_bytree' (0.8 and 1.0), 'gamma' (0 and 0.1), 'reg\_alpha' (0 and 0.1), and 'reg\_lambda' (0 and 0.1).
+    \item \textbf{Random Forests}: 
+    \begin{itemize}
+        \item \emph{Bootstrap}: True
+        \item \emph{Max\_features}: 'auto' and 'sqrt'
+        \item \emph{Max\_depth}: None, 10, and 20
+        \item \emph{Min\_samples\_leaf}: 1 and 2
+        \item \emph{Min\_samples\_split}: 2 and 5
+        \item \emph{N\_estimators}: 100, 200, and 300
+    \end{itemize}
+
+    \item \textbf{XGBoost}: 
+    \begin{itemize}
+        \item \emph{Max\_depth}: 3, 6, and 9
+        \item \emph{Learning\_rate}: 0.1 and 0.01
+        \item \emph{N\_estimators}: 100, 200, and 300
+        \item \emph{Subsample}: 0.8 and 1.0
+        \item \emph{Colsample\_bytree}: 0.8 and 1.0
+        \item \emph{Gamma}: 0 and 0.1
+        \item \emph{Reg\_alpha}: 0 and 0.1
+        \item \emph{Reg\_lambda}: 0 and 0.1
+    \end{itemize}
 \end{itemize}
@@
 
-\subsection{Artificial Neural Networks}
+\subsection{Neural Networks}
 
@@
 \begin{itemize}
-    \item 'learning\_rate' (0.001, 0.0001)
-    \item 'momentum' (0.5, 0.9)
-    \item 'number of neurons in hidden layer' (16, 32, 64)
-    \item 'dropout\_rate' (None, 0.2, 0.3)
-    \item 'batch\_size' (256, 512, 1028)
+    \item \emph{Learning Rate:} 0.001, 0.01, 0.1
+    \item \emph{Epochs:} 10, 50, 100
+    \item \emph{Batch Size:} 256, 512, 1028
+    \item \emph{Optimizer:} rmsprop, adam, SGD
+    \item \emph{Dropout Rate:} 0.2, 0.5
+    \item \emph{Number of Neurons in Hidden Layer(s):} 32, 64
+    \item \emph{Number of Hidden Layers:} 1, 3, 5
 \end{itemize}
 
-By evaluating these hyperparameters, we will to identify the optimal configuration of hyperparameters that maximizes the predictive performance of our ANN model on the given dataset.
-Additionally, we incorporate the Adam optimizer into our ANN model to further improve the training process and overall predictive accuracy.
+By evaluating these hyperparameters, we will to identify the optimal configuration of hyperparameters that maximizes the predictive performance of our neural network model on the given dataset.
+As shown above, we test the three different optimizers of our neural network model to achieve the best performance on this dataset.
 
@@
 In our grid search, we will use the \emph{negative mean squared error} as our metric to identify the optimal set of hyperparameters for both decision trees and the artificial neural network.
-To assess the performance of our models on the testing data, we will compare their results using several metrics, including Mean Squared Error (MSE), Mean Absolute Error (MAE), and R-squared ($R^2$) between the true values and our predictions.
+To assess the performance of our models on the testing data, we will compare their results using several metrics, including Mean Squared Error (MSE), Mean Absolute Error (MAE), and R-squared ($R^2$).
+Lastly, we'll also compute the Bayesian Information Criterion (BIC) score to assess model performance.
+Given the potential complexity of our decision tree and neural network models due to numerous parameters, the BIC score offers a balanced assessment by favoring models achieving a good fit while penalizing overly complex ones.
+This helps us ensure that the chosen model achieves both accuracy and parsimony, enhancing its usefulness for predicting diamond prices effectively.
+
 These metrics will help evaluating the accuracy and fit of each algorithm on the testing data.
@@
 
-Our experimental methodology provides nuanced results across each of the machine learning algorithms. In particular, the results are as follows:
+Our experimental methodology provides nuanced results across each of the machine learning algorithms. The results are as follows:
 
@@
 
-In terms of our training data, we tested the effectivness of both linear and polynomial regression models on our training data using 5-Fold Cross validation.
-Based on the results, this was our mean score error on training data:
+As mentioned previously, we employed a multiple linear regression model with all the features included, along with a polynomial regression model with the terms by raising the degree of features of our regression model.
+To find the optimal degree of the polynomial regression, we apply a 5-fold cross validation method on our dataset and plot its performance on both training and testing data.
+
+\begin{figure}[H]
+    \centering
+    \includegraphics[width=0.8\linewidth]{mae_poly.png} % Replace "example-image" with the filename of your image
+    \caption{MAE vs Degree of Polynomial Regression model}
+    \label{fig:boxplots}
+\end{figure}
+
+From our plot, we can see that the MAE on both training and testing data appears to be lowest when polynomial degree is squared.
+In terms of testing data, we display the performance scores of both algorithms on testing data, the results are displayed in the table below:
 
@@
 \centering
-\begin{tabular}{|c|c|c|}
+\caption{Linear vs.\ Polynomial Performance on Testing Data}
+\begin{tabular}{|c|c|c|c|}
 \hline
-Column 1 & Column 2 & Column 3 \\
+Method & MSE & MAE & $R^2$ \\
 \hline
-Row 1 & Cell 1 & Cell 2 \\
-Row 2 & Cell 3 & Cell 4 \\
-Row 3 & Cell 5 & Cell 6 \\
-Row 4 & Cell 7 & Cell 8 \\
+Linear Regression & 1110908.273 & 718.843 & 0.898 \\
+\hline
+Polynomial Regression &  459504.630 & 381.455 & 0.958 \\
 \hline
 \end{tabular}
-\caption{A simple table with 3 columns and 4 rows.}
 \label{tab:mytable}
@@
 
+In terms of the Mean Squared Error, the Polynomial Regression model outperforms the Linear Regression model, exhibiting a significantly lower MSE of 459,504.630 compared to 1,110,908.273 for Linear Regression.
+This suggests that the Polynomial Regression model provides more accurate predictions, with smaller errors on average.
+Similarly, the Mean Absolute Error (MAE) quantifies the average absolute difference between the predicted and actual diamond prices.
+Once again, the Polynomial Regression model shows superior performance, with a smaller MAE of 381.455 compared to 718.843 for Linear Regression.
+The coefficient of determination, $R^2$, represents the proportion of the variance in the diamond prices that is predictable from the independent variables.
+A value closer to 1 indicates a better fit of the model to the data.
+In this case, both models demonstrate strong performance, with $R^2$ values of 0.898 for Linear Regression and 0.958 for Polynomial Regression.
+However, the Polynomial Regression model exhibits a slightly higher $R^2$, indicating that it explains a greater proportion of the variance in diamond prices compared to the Linear Regression model.
+With this, our polynomial regression model exhibits a significant better performance on testing data with the quadratic relationship between predictors and price.
+Let's examine the performance for the more complex models, decision trees.
+
 \subsection{Decision Trees}
 
-\textbf{Random Forest Algorthm}
+First, we found the optimal results using 5-fold cross validation procedure for both random forest and XGBoost algorithm. The results from our Grid Search are as follows:
 
-For our decision tree algorirhtms, the results of grid searhc provided the following best set of hyperpatemrs for our random forest algorhtm.
-
-\begin{table}[h]
+\begin{table}[H]
 \centering
-\begin{tabular}{|c|c|c|}
+\caption{Optimal Hyperparameter for Random Forests}
+\begin{tabular}{|c|c|}
 \hline
-Column 1 & Column 2 & Column 3 \\
+Hyperparameter & Optimal Value \\
 \hline
-Row 1 & Cell 1 & Cell 2 \\
-Row 2 & Cell 3 & Cell 4 \\
-Row 3 & Cell 5 & Cell 6 \\
-Row 4 & Cell 7 & Cell 8 \\
+Bootstrap & TRUE \\
+\hline
+Max Depth & 20 \\
+\hline
+Max Features & 'sqrt' \\
+\hline
+Min Samples Leaf & 1 \\
+\hline
+Min Samples Split & 2 \\
+\hline
+N Estimators & 300 \\
 \hline
 \end{tabular}
-\caption{A simple table with 3 columns and 4 rows.}
 \label{tab:mytable}
@@
 
-\textbf{XGBoost}
+\emph{Random Forests} Ensemble of decision trees where each tree is built independently.
+Bootstrap sampling is used to create diverse datasets for training each tree.
+Trees are grown with a maximum depth of 20, allowing for relatively complex decision boundaries.
+Features are randomly sampled at each split, with the number of features chosen as the square root of the total features.
+A total of 300 trees are employed in the forest.
+The model can capture complex interactions between features due to deep trees and bootstrap sampling.
+Random feature selection adds further diversity, potentially reducing overfitting.
+High number of estimators provides robustness to noise in the data.
 
-\begin{table}[h]
+\begin{table}[H]
 \centering
-\begin{tabular}{|c|c|c|}
+\caption{Optimal Hyperparameter for XGBoost}
+\begin{tabular}{|c|c|}
 \hline
-Column 1 & Column 2 & Column 3 \\
+Hyperparameter & Optimal Value \\
 \hline
-Row 1 & Cell 1 & Cell 2 \\
-Row 2 & Cell 3 & Cell 4 \\
-Row 3 & Cell 5 & Cell 6 \\
-Row 4 & Cell 7 & Cell 8 \\
+Colsample\_byTree & 0.8 \\
+\hline
+Gamma & 0 \\
+\hline
+Learning Rate & 0.1 \\
+\hline
+Max Depth & 6 \\
+\hline
+Reg Alpha & 0 \\
+\hline
+N Estimators & 200 \\
+\hline
+Reg Lambda & 0.1 \\
+\hline
+Subsample & 1.0 \\
 \hline
 \end{tabular}
-\caption{A simple table with 3 columns and 4 rows.}
 \label{tab:mytable}
@@
 
+\emph{XGBoost} Gradient boosting ensemble method where each tree is built sequentially to correct the errors of the previous one.
+Features are sampled with a fraction of 0.8 for each tree, adding randomness and reducing overfitting.
+Trees are constrained to a maximum depth of 6, limiting the complexity of individual trees.
+Regularization terms, such as gamma and lambda, are set to 0 and 0.1 respectively, controlling overfitting.
+Utilizes a learning rate of 0.1 to shrink the contribution of each tree, enhancing model generalization.
+Employs 200 boosting rounds to iteratively improve model performance.
+Sequential training allows for the gradual refinement of predictions, focusing on the previously misclassified instances.
+Regularization terms and low learning rate prevent overfitting and ensure robustness.
+Limited tree depth balances model complexity and interpretability.
+
+\emph{Accuracy Scores} The table below displays accuracy measures on the testing data using the best hyperparameters.
+
+\begin{table}[H]
+\centering
+\caption{Linear vs.\ Polynomial Performance on Testing Data}
+\begin{tabular}{|c|c|c|c|}
+\hline
+Method & MSE & MAE & $R^2$ \\
+\hline
+Random Forests & 228,347.747 & 247.141 & 0.9803 \\
+\hline
+XGBoost & 195,937.71 & 233.941 & 0.9831 \\
+\hline
+\end{tabular}
+\label{tab:mytable}
+\end{table}
+
+Comparing the accuracy scores, we see that both Random Forests and XGBoost perform well, with XGBoost having slightly lower MSE and MAE, and a slightly higher $R^2$, indicating better performance overall.
+However, the difference is relatively small, and both models seem to be capturing the underlying patterns in the data effectively.
+They both perform better than regression methosd.
+
 \subsection{Neural Network}
 
-The grid search for neural network presents the following results:
+We used a 5-fold cross-validation procedure with an extensive Grid Search approach to find the optimal set of hyperparameters for our neural network. Here are the results:
 
-\begin{table}[h]
+\begin{table}[H]
 \centering
-\begin{tabular}{|c|c|c|}
+\caption{Optimal Hyperparameter for Random Forests}
+\begin{tabular}{|c|c|}
 \hline
-Column 1 & Column 2 & Column 3 \\
+Hyperparameter & Optimal Value \\
 \hline
-Row 1 & Cell 1 & Cell 2 \\
-Row 2 & Cell 3 & Cell 4 \\
-Row 3 & Cell 5 & Cell 6 \\
-Row 4 & Cell 7 & Cell 8 \\
+Learning Rate & XXX \\
+\hline
+Epochs & XXX \\
+\hline
+Batch Size & XXX \\
+\hline
+Optimizer & XXX \\
+\hline
+Dropout Rate & XXX \\
+\hline
+Number of Neurons in Hidden Layer & XXX \\
+\hline
+Number of Hidden Layers & XXX \\
 \hline
 \end{tabular}
-\caption{A simple table with 3 columns and 4 rows.}
+\label{tab:mytable}
+\end{table}
+
+The grid search method results in the architecture of our neural network model.
+Specifically, [...] Here, we present the training versus testing loss (MSE) across epochs during the training of our neural network model, along with its final performance on the testing data.
+
+\begin{figure}[H]
+    \centering
+    \includegraphics[width=0.8\linewidth]{train vs test.png} % Replace "example-image" with the filename of your image
+    \caption{Training vs Testing MSE Across Epoch}
+    \label{fig:boxplots}
+\end{figure}
+
+As we can see, based on our neural network architecture, the training and testing loss appear to go down quite a bit once we finish implementing across epochs. Thus this approach is good [...]
+
+\begin{table}[H]
+\centering
+\caption{Neural Network Performance on Testing Data}
+\begin{tabular}{|c|c|c|c|}
+\hline
+Method & MSE & MAE & $R^2$ \\
+\hline
+Neural Network & XXX & XXX & XXX \\
+\hline
+\end{tabular}
 \label{tab:mytable}
@@
 \subsection{Comparison of models on testing data}
+
+The table below displays the performance of each algorithm we've taken from above on the testing data along with its BIC criteria:
+
+\begin{table}[H]
+\centering
+\caption{Comparison of Algorithms}
+\begin{tabular}{|c|c|c|c|c|}
+\hline
+Method & MSE & MAE & $R^2$ & BIC \\
+\hline
+Linear Regression & 1,110,908.273 & 718.843 & 0.898 & XXX \\
+\hline
+Polynomial Regression & 459,504.630 & 381.455 &  0.958 & XXX \\
+\hline
+Random Forests & 228,347.747 & 247.141 & 0.9803 & XXX \\
+\hline
+XGBoost & 195,937.71 & 233.941 & 0.9831 & XXX \\
+\hline
+Neural Networks & XXX & XXX & XXX & XXX \\
+\hline
+\end{tabular}
+\label{tab:mytable}
+\end{table}
+
+[ADD ANALYSIS]
 
@@
 
+\section{Project RoadMap}
+
+Our project involves a regression-supervised learning task in the context of predicting the price of a diamond based on several different measures.
+Our dataset contains 53,940 observations and 9 columns (8 predictors and 1 response).
+The output variable “price” is continuous and has a range between $326 to $18,823.
+With this, our goal for this project is to create a machine-learning model using deep neural networks that can predict diamond prices as accurately as possible.
+The practical application of this project is quite significant, as an effective model can assist researchers and industry professionals in determining the value of diamonds in the market.
+As a part of developing our machine learning pipeline, we will first begin with exploratory data analysis.
+We do not have any missing values present, so we don’t have to worry about any data imputation for this specific project.
+First, we will begin by plotting the distribution of our features to understand the spread of our data and account for potential skewness.
+Since outliers can have significant effects on the overall accuracy of machine learning models, we will look for them in our dataset and remove them.
+We will use various outlier detection techniques such as one-class SVM and through the IQR computation; we're still planning the specific details as to how we will approach this step.
+Furthermore, since we have 3 categorical variables, we will use one-hot encoding and label encoding (in the case of an ordinal column) to convert them into numeric columns.
+Ultimately, our goal with EDA is to clean the dataset and understand the relationship between the predictors and the response variable.
+In terms of timeline, we plan to dedicate the EDA portion of our dataset this week/early next week (~ 1/1.5 weeks) before we begin training the model.
+We aim to have this portion of the project completed before the quarter midpoint and begin the process of training the model as well.
+
+Once EDA is complete, we are going to use deep neural networks to train a machine learning algorithm that will most accurately predict diamond prices.
+Neural networks are quite complex and include many hyper-parameters, so we want to explore several of them and see how they can help improve our results.
+In particular, we will begin by using K-fold cross-validation to train several neural networks of different sizes (with the number of layers and neurons varying) and compare the performance of each method in predicting diamond price.
+We will use the metrics learned in class, using total error and MSE as benchmarks for overall model performance.
+Furthermore, we are interested in exploring regularization techniques such as L2 regularization (possibly dropout/weight-decay on the learning rate as well, if time permits) to reduce overfitting and further improve the total error.
+Our goal is to produce the best model in terms of accuracy while also achieving a low training time; thus we will explore the implementation of different optimization techniques such as momentum SGD/RMSProp/Adam (choose one of three, likely momentum GD since we will be able to explain the concepts behind it clearly) to see how they affect our results in comparison to gradient descent.
+The details and concept of our optimization techniques will be included as part of the literature review section in the project report.
+Lastly, we may explore the effect of different activation functions (sigmoid/tanh) to see how the model performs in terms of accuracy and performance when non-linearity is introduced.
+Ultimately, our final goal is to develop a model such that we can get the best prediction accuracy on the generalized unseen data.
+
+In conclusion, this project not only aims to give a precise price prediction but also to provide insight into what important factors influence diamond prices—specific factors that may be of high value to a consumer and seller of diamonds.
+Rigorous data analysis and model-tuning efforts will be put in place to realize a robust model that can help demystify the complexities behind diamond pricing.
+In terms of the specific methodologies learned in our class, this project will help us understand how changes in the neural network architecture and tuning of hyper-parameters can affect our overall results, including changes in key metrics such as model accuracy and overall training speed.
+
+\emph{Group Contributions} The group had equal contributions throughout the course of this project.
+Andrew, Yifan, and Aditya focused on the model training and writing the paper.
+Nandini and Eshan worked on exploratory data analysis, setting up the Github repo, and creating the Project GUI environment for users.
+
+The \textbf{Github repository} for our project may be found here: https://github.com/adimittal03/ECS171-Final-Group21.
+
+\newpage
 \begin{thebibliography}{00}
@@
 \bibitem{kaggle} Agrawal, S. (2017a, May 25). Diamonds. Kaggle. https://www.kaggle.com/datasets/shivam2503/diamonds
+\bibitem{Bottou} Bottou, L. (2010). Large-Scale Machine Learning with Stochastic Gradient Descent. In Proceedings of COMPSTAT'2010, Springer. DOI: 10.1007/978-3-7908-2604-3\_16
+\bibitem{Kingma} Kingma, D. P., \& Ba, J. (2015). Adam: A Method for Stochastic Optimization. In International Conference on Learning Representations (ICLR). URL: https://arxiv.org/abs/1412.6980
+\bibitem{Tieleman} Tieleman, T., \& Hinton, G.
+(2012).
+Lecture 6.5 - RMSProp: Divide the gradient by a running average of its recent magnitude.
+COURSERA: Neural Networks for Machine Learning.
+URL: https://www.cs.toronto.edu/~tijmen/csc321/slides/lecture\_slides\_lec6.pdf
 \end{thebibliography}

### 2024-06-03 code/exploratory_data_analysis.py
includes all code: eda & model training
models tested: linear regression, polynomial regression (degree 2), random forests, XGBoost, Artificial Neural Network

@@ -1,152 +1,0 @@
-# import necessary modules
-import pandas as pd
-import numpy as np
-import matplotlib.pyplot as plt
-import seaborn as sns
-
-from sklearn.linear_model import LinearRegression
-from sklearn.preprocessing import PolynomialFeatures
-from sklearn.preprocessing import StandardScaler
-from sklearn.model_selection import train_test_split, GridSearchCV, KFold, cross_val_score
-from sklearn.metrics import mean_squared_error, r2_score
-from sklearn.ensemble import RandomForestRegressor
-import xgboost as xgb
-
-import tensorflow as tf
-from tensorflow.keras.models import Sequential
-from tensorflow.keras.layers import Dense, Dropout
-from tensorflow.keras import regularizers
-from tensorflow.keras.optimizers.legacy import Adam
-from tensorflow.keras.wrappers.scikit_learn import KerasRegressor
-
-### Exploratory Data Analysis
-### ----------------------------------------------------------------------------
-df = pd.read_csv("~/Desktop/Year_three/Spring_2024/ECS171/ECS171_Final/diamonds.csv") 
-df = df.iloc[:, 1:]  # remove index column          
-# df.head()
-
-### Plot distribution of continous features
-features_to_plot = ['carat', 'depth', 'table', 'x','y','z']
-plt.figure(figsize=(8, 6))
-for i, feature in enumerate(features_to_plot, 1):
-    plt.subplot(2, 3, i)
-    sns.histplot(df[feature], kde=True)
-    plt.title(feature)
-
-plt.tight_layout()
-plt.show()
-print("\nmean:\n{} , \nmedian:\n{} , \nstd:\n {}".format(df[features_to_plot].mean() , 
-                                                         df[features_to_plot].median(), 
-                                                         df[features_to_plot].std()))
-
-### Plot scatterplot of continous features vs. price
-feature_names = ['carat', 'depth', 'table', 'x','y','z']
-y = df['price']
-plt.figure(figsize=(15, 10))
-for i, feature in enumerate(feature_names):
-    plt.subplot(3, 2, i+1)
[…]

### 2024-06-03 code/project_code.py

@@ -1,0 +1,419 @@
+# import necessary modules
+import pandas as pd
+import numpy as np
+import matplotlib.pyplot as plt
+import seaborn as sns
+
+from sklearn.linear_model import LinearRegression
+from sklearn.preprocessing import PolynomialFeatures
+from sklearn.preprocessing import StandardScaler
+from sklearn.model_selection import train_test_split, GridSearchCV, KFold, cross_val_score
+from sklearn.metrics import mean_squared_error, r2_score, mean_absolute_error
+from sklearn.ensemble import RandomForestRegressor
+import xgboost as xgb
+
+import tensorflow as tf
+from tensorflow.keras.models import Sequential
+from tensorflow.keras.layers import Dense, Dropout
+from tensorflow.keras import regularizers
+from tensorflow.keras.optimizers.legacy import Adam, SGD, RMSprop
+from tensorflow.keras.wrappers.scikit_learn import KerasRegressor
+from tensorflow.keras import layers, models, optimizers
+from tensorflow import keras
+
+import joblib
+
+### Exploratory Data Analysis
+### ---------------------------------------------------------------------------------
+df = pd.read_csv("~/Desktop/Year_three/Spring_2024/ECS171/ECS171_Final/diamonds.csv") 
+df = df.iloc[:, 1:]  # remove index column          
+# df.head()
+
+### Plot distribution of continous features
+features_to_plot = ['carat', 'depth', 'table', 'x','y','z']
+plt.figure(figsize=(8, 6))
+for i, feature in enumerate(features_to_plot, 1):
+    plt.subplot(2, 3, i)
+    sns.histplot(df[feature], kde=True)
+    plt.title(feature)
+
+plt.tight_layout()
+plt.show()
+print("\nmean:\n{} , \nmedian:\n{} , \nstd:\n {}".format(df[features_to_plot].mean() , 
+                                                         df[features_to_plot].median(), 
+                                                         df[features_to_plot].std()))
+
+### Plot scatterplot of continous features vs. price
+feature_names = ['carat', 'depth', 'table', 'x','y','z']
+y = df['price']
+plt.figure(figsize=(15, 10))
[…]

### 2024-06-03 code/decision_trees.py
Delete code/decision_trees.py

@@ -1,87 +1,0 @@
-### Assuming we have X, X_test, y, y_test from the exploratory_data_analysis.py
-
-### Random Forests
-### ----------------------------------------------------------------------------
-### grid search
-rf = RandomForestRegressor()  
-
-rf_param_grid = {
-    'bootstrap': [True],
-    'max_features': ['auto', 'sqrt'],
-    'max_depth': [None, 10, 20],
-    'min_samples_leaf': [1, 2],
-    'min_samples_split': [2, 5],
-    'n_estimators': [100, 200, 300]
-}
-
-grid_search = GridSearchCV(estimator=rf, param_grid=rf_param_grid, cv=5, n_jobs=-1, verbose=0, scoring='neg_mean_squared_error')
-grid_search.fit(X, y)  
-
-print("Best Parameters:", grid_search.best_params_)
-print("Best Score:", grid_search.best_score_)
-
-### model fit on testing data
-best_params = {
-    'bootstrap': True,
-    'max_depth': 20,
-    'max_features': 'sqrt',
-    'min_samples_leaf': 1,
-    'min_samples_split': 2,
-    'n_estimators': 300
-}
-
-rf_regressor_best = RandomForestRegressor(**best_params, random_state=42)
-rf_regressor_best.fit(X, y)
-
-y_pred = rf_regressor_best.predict(X_test)
-mae = np.mean(abs(y_pred - y_test))
-r_squared = r2_score(y_test, y_pred)
-mse = mean_squared_error(y_test, y_pred)
-print(f'MAE: {mae}')
-print("R-squared:", r_squared)
-print("Mean Squared Error:", mse)
-
-### XGBoost
-### ----------------------------------------------------------------------------
-### grid search
-
-xgb_model = xgb.XGBRegressor()  
-xgb_param_grid = {
-    'max_depth': [3, 6, 9],  
-    'learning_rate': [0.1, 0.01],  
-    'n_estimators': [100, 200, 300],  
-    'subsample': [0.8, 1.0],  
-    'colsample_bytree': [0.8, 1.0],  
-    'gamma': [0, 0.1],  
-    'reg_alpha': [0, 0.1],  
-    'reg_lambda': [0, 0.1]  
-}
-grid_search_xgb = GridSearchCV(estimator=xgb_model, param_grid=xgb_param_grid, cv=5, scoring='neg_mean_squared_error', verbose=0)
-grid_search_xgb.fit(X, y)  
[…]

### 2024-06-03 code/neural_network.py
Delete code/neural_network.py

@@ -1,56 +1,0 @@
-### Assuming we have X, X_test, y, y_test from exploratory_data_analysis.py
-
-### Defining the Neural Network ------------------------------------------------
-### Grid Search to find best hyper-params
-def create_model(num_layers = 3, num_neurons = 64, dropout_rate=0.0, optimizer='adam', learning_rate = 0.01):
-    model = Sequential()
-
-    # hidden layer one
-    model.add(Dense(num_neurons, input_dim=9, activation='relu'))  
-    model.add(Dropout(dropout_rate))
-    
-    # following hidden layers
-    for _ in range(num_layers - 1):
-        model.add(Dense(num_neurons, activation='relu'))
-        model.add(Dropout(dropout_rate))
-    
-    # output layer
-    model.add(Dense(1, activation='relu'))
-    
-    # optimizer
-    if optimizer == 'rmsprop':
-        opt = tf.keras.optimizers.legacy.RMSprop(learning_rate=learning_rate)
-    elif optimizer == 'adam':
-        opt = tf.keras.optimizers.legacy.Adam(learning_rate=learning_rate)
-    elif optimizer == 'SGD':
-        opt = tf.keras.optimizers.legacy.SGD(learning_rate=learning_rate)
-    else:
-        raise ValueError('optimizer {} unrecognized'.format(optimizer))
-        
-    # compile
-    network.compile(optimizer = opt, 
-                loss = 'mean_squared_error', 
-                metrics=['mae', tf.keras.metrics.RootMeanSquaredError()])
-    
-    return model
-
-### Implement Grid Search -------------------------------------------------------
-ann = KerasRegressor(build_fn=create_model, verbose=0)
-
-# these are the values for the other hyperparameters
-param_grid = dict(
-    num_layers= [1, 3, 5, 8],
-    num_neurons= [16, 32, 64],
-    dropout_rate= [0.2, 0.5],
-    optimizer = ['rmsprop', 'adam', 'SGD'],
-    batch_size = [32, 64, 128, 256, 512, 1024, 2048],
-    epochs = [10, 25, 50, 100, 150, 200],
-    learning_rate = [0.001, 0.01, 0.1, 0.2, 0.3]
-)
-
-# apply 5-fold CV to grid search
-grid = GridSearchCV(estimator=ann, cv=5, param_grid=param_grid)
-grid_result = grid.fit(X, y)
-
-# print best set of parameters // comes after the warnings 
-print("Best: %f using %s" % (grid_result.best_score_, grid_result.best_params_))

### 2024-06-03 code/regression_models.py
Delete code/regression_models.py

@@ -1,55 +1,0 @@
-### Assuming we have X, X_test, y, y_test from exploratory_data_analysis.py
-
-### Linear Regression
-### ----------------------------------------------------------------------------
-
-### Cross validation procedure
-model = LinearRegression()
-kf = KFold(n_splits=5, shuffle=True, random_state=42) 
-
-mse_scores = cross_val_score(model, X, y, cv=kf, scoring='neg_mean_squared_error')
-mse_scores = -mse_scores
-mean_mse = np.mean(mse_scores)
-std_mse = np.std(mse_scores)
-print(f'Mean MSE: {mean_mse}')
-print(f'Standard Deviation of MSE: {std_mse}')
-
-
-### Model accuracy on testing data
-model = LinearRegression()
-model.fit(X, y)
-y_pred = model.predict(X_test)
-mae = np.mean(abs(y_pred - y_test))
-r_squared = r2_score(y_test, y_pred)
-mse = mean_squared_error(y_test, y_pred)
-print(f'MAE: {mae}')
-print("R-squared:", r_squared)
-print("Mean Squared Error:", mse)
-
-
-### Polynomial Regression
-### ----------------------------------------------------------------------------
-poly = PolynomialFeatures(2)
-X_poly = poly.fit_transform(X)
-X_poly_test = poly.fit_transform(X_test)
-
-# Cross validation procedure
-model = LinearRegression()
-kf = KFold(n_splits=10, shuffle=True, random_state=42) # 5 fold CV
-mse_scores = cross_val_score(model, X_poly, y, cv=kf, scoring='neg_mean_squared_error')
-mse_scores = -mse_scores
-mean_mse = np.mean(mse_scores)
-std_mse = np.std(mse_scores)
-print(f'Mean MSE: {mean_mse}')
-print(f'Standard Deviation of MSE: {std_mse}')
-
-# model fit on testing data
-model = LinearRegression()
-model.fit(X_poly, y)
-y_pred = model.predict(X_poly_test)
-mae = np.mean(abs(y_pred - y_test))
-r_squared = r2_score(y_test, y_pred)
-mse = mean_squared_error(y_test, y_pred)
-print(f'MAE: {mae}')
-print("R-squared:", r_squared)
-print("Mean Squared Error:", mse)

### 2024-06-03 flask_gui/app.py
create app.py for flask

@@ -1,0 +1,88 @@
+from flask import Flask, render_template, request
+import joblib
+import numpy as np
+import tensorflow as tf
+from tensorflow.keras.models import model_from_json
+from sklearn.preprocessing import PolynomialFeatures
+from sklearn.model_selection import KFold, cross_val_score
+from sklearn.linear_model import LinearRegression
+from sklearn.metrics import mean_absolute_error, r2_score, mean_squared_error
+
+app = Flask(__name__)
+
+# Load your trained models and scaler
+models = {
+    'linear_regression': joblib.load('linear_regression_model.joblib'),
+    'random_forest': joblib.load('random_forest_regressor_model.joblib'),
+    'XGBoost': joblib.load('xgboost_regressor_model.joblib')
+}
+
+# Load the polynomial regression model
+poly_model = joblib.load('polynomial_regression_model.joblib')
+
+# Load the TensorFlow model architecture from the JSON file
+try:
+    with open("model.json", "r") as json_file:
+        loaded_model_json = json_file.read()
+        neural_network_model = model_from_json(loaded_model_json)
+    neural_network_model.load_weights("model_weights.h5")
+except Exception as e:
+    print("Error loading TensorFlow model:", e)
+    neural_network_model = None
+
+# Load the scaler
+scaler = joblib.load('scaler.joblib')
+
+@app.route('/')
+def home():
+    return render_template('index.html')
+
+@app.route('/result', methods=['POST'])
+def predict():
+    try:
+        # Get the input values from the form
+        carat = float(request.form['carat'])
+        depth = float(request.form['depth'])
+        table = float(request.form['table'])
+        x = float(request.form['x'])
+        y = float(request.form['y'])
+        z = float(request.form['z'])
+        cut_encoded = int(request.form['cut_encoded'])
+        color_encoded = int(request.form['color_encoded'])
+        clarity_encoded = int(request.form['clarity_encoded'])
+        
+        # Prepare input data for prediction
+        input_data = np.array([[carat, depth, table, x, y, z, cut_encoded, color_encoded, clarity_encoded]])
+        
+        # Preprocess input data for polynomial regression model
+        poly = PolynomialFeatures(2)
+        input_data_poly = poly.fit_transform(input_data)
+        
+        # Scale the input data
+        input_data_scaled = scaler.transform(input_data)
+        
+        # Make predictions using all models
+        predictions = {}
+        for model_name, model in models.items():
+            prediction = model.predict(input_data_scaled)
+            predictions[model_name] = prediction[0]
+        
[…]

### 2024-06-03 paper/main.tex
neural network example

@@
 \usepackage{textcomp}
+%\usepackage{subcaption}
 \usepackage{natbib}
@@
 Davis, United States \\
-XXX}
+harcui@ucdavis.edu}
 \and
@@
     \centering
-    \includegraphics[width=0.745\linewidth]{Figure_1.png}
+    \includegraphics[width=0.7\linewidth]{Figure_1.png}
     \caption{Distribution Plot of Continuous Features. Identify trends in our dataset such as skewness, location of outliers, mean, median, scale, etc.}
@@
     \centering
-    \includegraphics[width=0.8\linewidth]{Figure_4.png} % Replace "example-image" with the filename of your image
+    \includegraphics[width=0.7\linewidth]{Figure_4.png} % Replace "example-image" with the filename of your image
     \caption{Pair-wise Scatterplots for Each Feature vs. Diamond Price}
@@
     \centering
-    \includegraphics[width=0.8\linewidth]{Figure_3.png} % Replace "example-image" with the filename of your image
+    \includegraphics[width=0.7\linewidth]{Figure_3.png} % Replace "example-image" with the filename of your image
     \caption{Correlation Heat-Map for Continuous Features vs Diamond Price}
@@
 
-\subsection{Linear Regression}
+\subsection{Regression}
 
@@
 
-In our exploration of predicting diamond prices, we employ artificial neural networks (ANNs).
-Our ANN architecture consists of a single hidden layer with ReLU activation, followed by an output layer also utilizing ReLU activation, as our task involves regression.
-To optimize the performance of our ANN, we conduct a grid search for hyperparameter tuning, aiming to select the most suitable parameters for our model.
+In our exploration of predicting diamond prices, we employ neural networks.
+Our neural network architecture consists of a single hidden layer with ReLU activation, followed by an output layer also utilizing ReLU activation, as our task involves regression.
+To optimize the performance of our model, we conduct a grid search for hyperparameter tuning, aiming to select the most suitable parameters for our model.
 We explore the following hyperparameters during grid search with a 5-fold cross-validation procedure.
@@
 \hline
-Learning Rate & XXX \\
+Learning Rate & 0.1 \\
 \hline
-Epochs & XXX \\
+Epochs & 250 \\
 \hline
-Batch Size & XXX \\
+Batch Size & 256 \\
 \hline
-Optimizer & XXX \\
+Optimizer & RMSProp \\
 \hline
-Dropout Rate & XXX \\
+Dropout Rate & None \\
 \hline
-Number of Neurons in Hidden Layer & XXX \\
-\hline
-Number of Hidden Layers & XXX \\
+Number of Neurons in Hidden Layer & 64 \\
 \hline
@@
 The grid search method results in the architecture of our neural network model.
-Specifically, [...] Here, we present the training versus testing loss (MSE) across epochs during the training of our neural network model, along with its final performance on the testing data.
+Specifically, the neural network architecture seems to consist of a single hidden layer with 64 neurons, employing a learning rate of 0.1, RMSProp optimizer, and no dropout regularization.
+It underwent training for 250 epochs with a batch size of 256.
+Here, we present the training versus testing loss (MSE) across epochs during the training of our neural network model, along with its final performance on the testing data.
 
@@
     \centering
-    \includegraphics[width=0.8\linewidth]{train vs test.png} % Replace "example-image" with the filename of your image
-    \caption{Training vs Testing MSE Across Epoch}
+    \includegraphics[width=0.8\linewidth]{1000.png} % Replace "example-image" with the filename of your image
+    \caption{MAE vs Degree of Polynomial Regression model}
+    \label{fig:boxplots}
+\end{figure}
+
+From our results of training the neural network across 1000 epochs, we observe that the model effectively learns the weights.
+Specifically, both the training and testing errors significantly reduce during the initial stages of training.
+However, around 500 epochs, our neural network model begins to overfit the data: the testing error starts to increase significantly while the training error continues to decrease.
+This indicates that we can use 250 epochs as our stopping point for neural network training to avoid overfitting.
+
+\begin{figure}[H]
+    \centering
+    \includegraphics[width=0.8\linewidth]{250.png} % Replace "example-image" with the filename of your image
+    \caption{MAE vs Degree of Polynomial Regression model}
     \label{fig:boxplots}
@@
 \hline
-Neural Network & XXX & XXX & XXX \\
+Neural Network & 382,237.13 & 272.47 & 0.9652  \\
 \hline
@@
 \hline
-Method & MSE & MAE & $R^2$ & BIC \\
+Method & MSE & MAE & $R^2$ \\
 \hline
-Linear Regression & 1,110,908.273 & 718.843 & 0.898 & XXX \\
+Linear Regression & 1,110,908.273 & 718.843 & 0.898 \\
 \hline
-Polynomial Regression & 459,504.630 & 381.455 &  0.958 & XXX \\
+Polynomial Regression & 459,504.630 & 381.455 &  0.958 \\
 \hline
-Random Forests & 228,347.747 & 247.141 & 0.9803 & XXX \\
+Random Forests & 228,347.747 & 247.141 & 0.9803  \\
 \hline
-XGBoost & 195,937.71 & 233.941 & 0.9831 & XXX \\
+XGBoost & 195,937.71 & 233.941 & 0.9831 \\
 \hline
-Neural Networks & XXX & XXX & XXX & XXX \\
+Neural Networks & 382,237.13 & 272.47 & 0.9652 \\
 \hline
@@
 
-[ADD ANALYSIS]
+The comparison of algorithms for diamond pricing reveals that XGBoost performs the best, achieving the lowest MSE (195,937.71) and MAE (233.941), along with the highest $R^2$ (0.9831), indicating it provides the most accurate price predictions and best explains the variance in diamond prices.
+Random Forests also demonstrate strong performance, closely following XGBoost.
+Neural Networks and Polynomial Regression are effective but exhibit higher error rates and slightly lower $R^2$ values, making them less precise.
+Linear Regression, though straightforward and easy to interpret, performs the worst with the highest MSE and MAE, and the lowest $R^2$, suggesting it is the least reliable for predicting diamond prices.
+Therefore, ensemble methods like XGBoost and Random Forests are the most effective models for accurate diamond pricing, while Linear Regression is the least suitable.
 
@@
 
-\section{Project RoadMap}
+\section{Project Road-Map}
 
@@
 Our dataset contains 53,940 observations and 9 columns (8 predictors and 1 response).
-The output variable “price” is continuous and has a range between $326 to $18,823.
+The output variable “price” is continuous and has a range between \$326 to \$18,823.
 With this, our goal for this project is to create a machine-learning model using deep neural networks that can predict diamond prices as accurately as possible.
@@
 URL: https://www.cs.toronto.edu/~tijmen/csc321/slides/lecture\_slides\_lec6.pdf
+
 \end{thebibliography}

### 2024-06-04 paper/main.tex
Update main.tex

@@
     T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
-\begin{document}
 
 \title{Machine Learning Approaches for Accurate Diamond Price Prediction\\
-% {\footnotesize ECS-171 Final Project: \textbf{Group 21}}
 \thanks{Final Project Report ECS-171: \textbf{Group 21}}
@@
 
+\begin{document} 
 \maketitle
@@
 jewelry market was valued at 73.32 billion USD, with diamonds comprising a significant share.
-Predicting diamond prices accurately is essential for jewelers to make informed pricing decisions.
-This paper explores various machine learning algorithms, including linear regression, Random Forests, XGBoost, and neural networks, to predict diamond prices based on characteristics such as carat, cut, color, and clarity.
-Through exploratory data analysis, we identify key features influencing diamond prices and apply these insights to improve model performance.
-Our methodology includes rigorous model evaluation using metrics like Mean Squared Error (MSE), Mean Absolute Error (MAE), and R-squared (R²).
+Accurate diamond price prediction is essential for jewelers.
+This paper explores machine learning algorithms, including linear regression, polynomial regression, Random Forests, XGBoost, and neural networks, to predict diamond prices based on carat, cut, color, and clarity.
+Exploratory data analysis identified carat as the most influential feature.
+Our methodology involves rigorous model evaluation using metrics like Mean Squared Error (MSE), Mean Absolute Error (MAE), and R-squared (R²).
+Experimental results show XGBoost outperformed other models, achieving the lowest testing error, followed closely by Random Forests.
+Polynomial regression also performed well, highlighting the importance of capturing non-linear relationships.
+The neural network model, while promising, needs further optimization.
 By developing a robust predictive model, we aim to enhance pricing precision, benefiting both businesses and consumers in the diamond market.
@@
     \centering
-    \includegraphics[width=0.7\linewidth]{Figure_1.png}
+    \includegraphics[width=0.53\linewidth]{Figure_1.png}
     \caption{Distribution Plot of Continuous Features. Identify trends in our dataset such as skewness, location of outliers, mean, median, scale, etc.}
@@
     \caption{Pair-wise Scatterplots for Each Feature vs. Diamond Price}
-    \label{fig:image_label}
+    \label{fig:scatter}
 \end{figure}
@@
 
-In this paper, we will explore the predictive capabilities of several machine learning algorithms, ranging from linear regression, decision trees, and artificial neural networks.
-Our approach involved several steps to develop and evaluate model performance to choose the most accurate algorithm.
-Data pre-processing has already been completed as described above.
-We divided the dataset into training and testing sets using a 75:25 split.
-The training data consists of 38,000 rows, while the testing data contains 12,000 rows.
-We use cross validation procedures (detailed below) to find the optimal degree of our polynomial regression model and set of optimal hyperparameters for the decision trees \& neural network models.
+In this paper, we will implement various machine learning algorithms to generate accurate diamond predictions.
+The models will include linear regression, decision trees, and artificial neural networks.
+Our approach involves several steps to develop and evaluate model performance to choose the best algorithm for this application.
+Data preprocessing has been completed as described earlier.
+The dataset has been divided into training and testing sets using a 75:25 split.
+The training data comprises 38,200 rows, while the testing data contains approximately 12,800 rows.
+We apply cross-validation procedures (detailed below) on our training data to determine the optimal degree of our polynomial regression model and the optimal set of hyperparameters for the decision trees and neural network models.
 
-\subsection{Regression}
+\subsection{Regression Models}
 
-We begin with linear regression model to fit a straight-line model between diamond prices and the aforementioned features.
-This initial step helps us gain intuition about the relationship between the features and diamond prices using a simple model.
-From our EDA step, since price displayed a quadratic relationship with carat and diamond dimensions (x,y,z), we implement a polynomial regression model to test its performance.
-While the graph indicates a potential quadratic relationship, we employ 5-fold cross-validation by testing powers from 1 to find the optimal degree of our regression model.This allows us to capture nonlinear relationships and assess whether polynomial regression outperforms traditional linear regression.
+We begin with the simplest algorithm: the Linear regression model.
+This algorithm fits a straight-line model between diamond prices and the aforementioned features.
+This initial model will help us gain intuition about the relationship between the features and diamond prices.
+Additionally, since our Exploratory Data Analysis (EDA) results indicating a potential quadratic relationship between price and carat \& diamond dimensions (x, y, z), we also implement a Polynomial Regression model and test its performance.
+Using 5-fold cross-validation, we test powers ranging from 1 to 5 to determine the optimal degree of our regression model.
+This allows us to identify whether more complex relationships in our data exist, capture nonlinear relationships, and assess whether Polynomial Regression outperforms traditional Linear Regression.
+After determining the optimal degree of our polynomial regression model, we will train the algorithm on the entire training dataset.
+Finally, we evaluate its performance on the test set to compare it with other algorithms.
 
@@
 
-For decision tree-based models, we employ two algorithms: Random Forests and XGBoost.
+For decision tree-based models, we implement two algorithms: Random Forests and XGBoost.
 Random Forests aggregate the predictions of multiple decision trees to enhance accuracy and reduce overfitting; XGBoost is a gradient boosting algorithm that iteratively optimizes the model to minimize overall error.
-We implement grid search with a 5-fold cross-validation to identify the optimal set of hyperparameters.
-These are the following set of hyperparameters we tested:
+To determine the optimal set of hyperparameters for each algorithm, we employ grid search with 5-fold cross-validation on our training data.
+The following sets of hyperparameters were tested for each algorithm:
 
@@
         \item \emph{Max\_features}: 'auto' and 'sqrt'
-        \item \emph{Max\_depth}: None, 10, and 20
+        \item \emph{Max\_depth}: None, 5, and 10
         \item \emph{Min\_samples\_leaf}: 1 and 2
-        \item \emph{Min\_samples\_split}: 2 and 5
-        \item \emph{N\_estimators}: 100, 200, and 300
+        \item \emph{Min\_samples\_split}: 1 and 2
+        \item \emph{N\_estimators}: 1, 5, and 10
     \end{itemize}
@@
 
-This broad range of values for each hyperparameter enables us to explore a wide array of model configurations and identify the optimal combination for our dataset.
-Subsequently, we train the model using this set of hyperparameters on the training data and assess its performance on the testing set.
-The plot of training and testing loss vs.\ epoch is provided to demonstrate model performance.
+Using grid search, we explore various hyperparameter values to assess different model configurations and identify the optimal combination for our dataset.
+This comprehensive approach allows us to evaluate both simple and complex models, informing us about the algorithm configuration that yields the most accurate predictions.
+Subsequently, we train the model using these optimized hyperparameters on the training data and evaluate its performance on the testing set.
+We compare results with the regression and neural network models.
 
@@
 
-In our exploration of predicting diamond prices, we employ neural networks.
-Our neural network architecture consists of a single hidden layer with ReLU activation, followed by an output layer also utilizing ReLU activation, as our task involves regression.
-To optimize the performance of our model, we conduct a grid search for hyperparameter tuning, aiming to select the most suitable parameters for our model.
-We explore the following hyperparameters during grid search with a 5-fold cross-validation procedure.
+Our final model being testing is an artificial neural network.
+The architecture consists of a single hidden layer with ReLU activation, followed by an output layer also using ReLU activation, given our regression task.
+The hyperparameters explored during grid search are found using a 5-fold cross-validation procedure on our training data; these include learning rate, batch size, optimizer, dropout rate, and the number of neurons in the hidden layer.
+Through this process, we will identify the optimal configuration of hyperparameters that maximizes the predictive performance of our neural network model on the given dataset.
 We explore the following set of hyperparameters:
@@
     \item \emph{Learning Rate:} 0.001, 0.01, 0.1
-    \item \emph{Epochs:} 10, 50, 100
     \item \emph{Batch Size:} 256, 512, 1028
     \item \emph{Optimizer:} rmsprop, adam, SGD
-    \item \emph{Dropout Rate:} 0.2, 0.5
-    \item \emph{Number of Neurons in Hidden Layer(s):} 32, 64
-    \item \emph{Number of Hidden Layers:} 1, 3, 5
+    \item \emph{Dropout Rate:} 0, 0.2, 0.5
+    \item \emph{Number of Neurons in Hidden Layer(s):} 16, 32, 64
 \end{itemize}
 
-By evaluating these hyperparameters, we will to identify the optimal configuration of hyperparameters that maximizes the predictive performance of our neural network model on the given dataset.
-As shown above, we test the three different optimizers of our neural network model to achieve the best performance on this dataset.
+This comprehensive exploration of hyperparameters will provide insights into the set of hyperparameters that give the best results.
+We will evaluate the impact of different learning rates and batch sizes on the model convergence, while adjusting the number of hidden layers and dropout rates manage model complexity and implement potential regularization in the hidden layer.
+As shown above, we test the three different optimizers of our neural network model to achieve the best performance on this dataset with the corresponding hyperparameters.
+By evaluating these hyperparameters, our aim is to optimize the neural network model's performance and improve its predictive accuracy for the given diamond dataset.
 
@@
 
-In our grid search, we will use the \emph{negative mean squared error} as our metric to identify the optimal set of hyperparameters for both decision trees and the artificial neural network.
-To assess the performance of our models on the testing data, we will compare their results using several metrics, including Mean Squared Error (MSE), Mean Absolute Error (MAE), and R-squared ($R^2$).
-Lastly, we'll also compute the Bayesian Information Criterion (BIC) score to assess model performance.
-Given the potential complexity of our decision tree and neural network models due to numerous parameters, the BIC score offers a balanced assessment by favoring models achieving a good fit while penalizing overly complex ones.
-This helps us ensure that the chosen model achieves both accuracy and parsimony, enhancing its usefulness for predicting diamond prices effectively.
-
-These metrics will help evaluating the accuracy and fit of each algorithm on the testing data.
-By analyzing these measures, our goal is to find the most effective algorithm for predicting diamond prices as it offers valuable insights for jewelers and marketers in the industry.
-Through this methodology, we will develop an accurate for predicting diamond prices, leveraging the strengths of various machine learning techniques and achieve best performance possible.
+To identify the optimal hyperparameters in our grid search, we will use the \emph{negative mean squared error} as our evaluation metric for both decision trees and the artificial neural network.
+With this metric, we aim to determine the best set of hyperparameters for each model.
+We will subsequently train the models using these optimal hyperparameters on the entire training data.
+To assess model performance on the testing data and compare the algorithms, we will compare results using various metrics, including Mean Squared Error (MSE), Mean Absolute Error (MAE), and R-squared ($R^2$).
+These metrics offer a comprehensive understanding of model performance, aiding in evaluating accuracy and fit.
+Additionally, we will display the training time for each model, providing insight into computational complexity.
+This evaluation process aims to identify the most effective algorithm for predicting diamond prices, offering valuable insights for both jewelers and consumers in the industry.
+For jewelers, accurate price predictions enable better inventory management and pricing strategies, while consumers benefit from informed purchasing decisions.
+Through this methodology, we aim to develop an accurate model for predicting diamond prices, leveraging the strengths of various machine learning techniques to achieve the best possible performance.
 
@@
 
-Our experimental methodology provides nuanced results across each of the machine learning algorithms. The results are as follows:
+In this section, we present our experimental results where we evaluated the performance of each machine learning algorithm.
+Through this analysis, we aim to demonstrate the effectiveness and suitability of each approach, offering valuable insights for deploying commercial algorithms in predicting diamond prices.
 
@@
 
-As mentioned previously, we employed a multiple linear regression model with all the features included, along with a polynomial regression model with the terms by raising the degree of features of our regression model.
-To find the optimal degree of the polynomial regression, we apply a 5-fold cross validation method on our dataset and plot its performance on both training and testing data.
+As previously mentioned, we've implemented both multiple linear regression and polynomial regression models using all available features in our dataset.
+During the exploratory data analysis (EDA) phase, we observed a possible quadratic relationship between diamond prices and the predictors as shown in Figure \ref{fig:scatter}.
+However, more complex relationships may also exist.
+To determine the optimal degree for the polynomial regression model, we're using a 5-fold cross-validation technique on our training dataset.
+We'll plot its performance on both the training and testing data:
 
@@
     \includegraphics[width=0.8\linewidth]{mae_poly.png} % Replace "example-image" with the filename of your image
-    \caption{MAE vs Degree of Polynomial Regression model}
-    \label{fig:boxplots}
+    \caption{Mean Absolute Error (MAE) on Training/Testing Data vs Degree of Polynomial Regression Model. Lowest train/test error can be see at degree = 2.}
+    \label{fig:degree}
 \end{figure}
 
-From our plot, we can see that the MAE on both training and testing data appears to be lowest when polynomial degree is squared.
-In terms of testing data, we display the performance scores of both algorithms on testing data, the results are displayed in the table below:
+Based on our results in Figure \ref{fig:degree}, we observed that the Mean Absolute Error was the lowest on both training and testing data when the degree of the polynomial model was squared.
+As we increased the degree to 3, both training and testing errors seemed to rise, indicating that the model might not effectively capture the underlying relationship in this dataset.
+Thus, our results from cross validation support our findings from the EDA phase that a quadratic relationship is appropriate.
+We present the performance of our model on the training data from cross-validation, along with training time, below.
 
@@
 \centering
-\caption{Linear vs.\ Polynomial Performance on Testing Data}
+\caption{Linear vs.\ Polynomial Performance on Training Data}
 \begin{tabular}{|c|c|c|c|}
 \hline
-Method & MSE & MAE & $R^2$ \\
+Method & Mean MSE & Variance MSE & Time \\
 \hline
-Linear Regression & 1110908.273 & 718.843 & 0.898 \\
+Linear Regression & 1,129,207.421 & 47,218.77 & 0.0075(s) \\
 \hline
-Polynomial Regression &  459504.630 & 381.455 & 0.958 \\
+Polynomial Regression &  9,055,852.513 & 55,599.664 & 0.072 (s) \\
 \hline
 \end{tabular}
-\label{tab:mytable}
+\label{tab:lr_trn}
 \end{table}
 
-In terms of the Mean Squared Error, the Polynomial Regression model outperforms the Linear Regression model, exhibiting a significantly lower MSE of 459,504.630 compared to 1,110,908.273 for Linear Regression.
-This suggests that the Polynomial Regression model provides more accurate predictions, with smaller errors on average.
-Similarly, the Mean Absolute Error (MAE) quantifies the average absolute difference between the predicted and actual diamond prices.
-Once again, the Polynomial Regression model shows superior performance, with a smaller MAE of 381.455 compared to 718.843 for Linear Regression.
-The coefficient of determination, $R^2$, represents the proportion of the variance in the diamond prices that is predictable from the independent variables.
-A value closer to 1 indicates a better fit of the model to the data.
-In this case, both models demonstrate strong performance, with $R^2$ values of 0.898 for Linear Regression and 0.958 for Polynomial Regression.
-However, the Polynomial Regression model exhibits a slightly higher $R^2$, indicating that it explains a greater proportion of the variance in diamond prices compared to the Linear Regression model.
-With this, our polynomial regression model exhibits a significant better performance on testing data with the quadratic relationship between predictors and price.
-Let's examine the performance for the more complex models, decision trees.
+Our findings from model training reveal that polynomial regression, particularly with a squared degree, yields a much higher average training error (and variance) compared to the linear regression model across the 5-fold cross-validation.
+Additionally, polynomial regression incurs a significant increase in computational cost, with its training time nearly ten times longer than that of linear regression.
+Despite both algorithms having relatively quick training speeds due to their overall simplicity as compared to the other models, the substantial computational overhead of polynomial regression underscores the need to carefully weigh its benefits in capturing non-linear relationships against the associated computational demands.
+We will also evaluate its performance on the test set and compare with other models below.
 
@@
 
-First, we found the optimal results using 5-fold cross validation procedure for both random forest and XGBoost algorithm. The results from our Grid Search are as follows:
+To determine the optimal set of hyperparameters, we conducted a 5-fold cross-validation procedure on our training data for both the Random Forest and XGBoost models.
+The table below presents the optimal hyperparameter settings for the Random Forest algorithm, selected from the parameter grid proposed earlier:
 
@@
 \hline
-Max Depth & 20 \\
+Max Depth & None \\
 \hline
@@
 \hline
-Min Samples Leaf & 1 \\
+Min Samples Leaf & 2 \\
 \hline
@@
 \hline
-N Estimators & 300 \\
+N Estimators & 10 \\
 \hline
 \end{tabular}
-\label{tab:mytable}
+\label{tab:rf_hyper}
 \end{table}
 
-\emph{Random Forests} Ensemble of decision trees where each tree is built independently.
-Bootstrap sampling is used to create diverse datasets for training each tree.
-Trees are grown with a maximum depth of 20, allowing for relatively complex decision boundaries.
-Features are randomly sampled at each split, with the number of features chosen as the square root of the total features.
-A total of 300 trees are employed in the forest.
-The model can capture complex interactions between features due to deep trees and bootstrap sampling.
-Random feature selection adds further diversity, potentially reducing overfitting.
-High number of estimators provides robustness to noise in the data.
+\emph{Random Forests} displays the optimal hyperparameters, providing insight into the Random Forest algorithm’s structure (i) Bootstrap sampling ('Bootstrap = TRUE') diversifies training datasets for each tree.
+(ii) The absence of a maximum depth ('Max Depth = None') allows trees to grow deep and simplifies decision boundaries.
+(iii) Features are randomly sampled at each split ('Max Features = sqrt').
+(iv) Both leaf and internal nodes require a minimum of 2 samples ('Min Samples Leaf = 2' and 'Min Samples Split = 2').
+(v) Utilizing 10 trees ('N Estimators = 10') enables capturing feature interactions.
+This combination of hyperparameters should enable the Random Forest algorithm to achieve accurate generalization while also potentially mitigating overfitting.
 
@@
 \end{tabular}
-\label{tab:mytable}
+\label{tab:xg_hyper}
 \end{table}
 
-\emph{XGBoost} Gradient boosting ensemble method where each tree is built sequentially to correct the errors of the previous one.
-Features are sampled with a fraction of 0.8 for each tree, adding randomness and reducing overfitting.
-Trees are constrained to a maximum depth of 6, limiting the complexity of individual trees.
-Regularization terms, such as gamma and lambda, are set to 0 and 0.1 respectively, controlling overfitting.
-Utilizes a learning rate of 0.1 to shrink the contribution of each tree, enhancing model generalization.
-Employs 200 boosting rounds to iteratively improve model performance.
-Sequential training allows for the gradual refinement of predictions, focusing on the previously misclassified instances.
-Regularization terms and low learning rate prevent overfitting and ensure robustness.
-Limited tree depth balances model complexity and interpretability.
+\emph{XGBoost}: Table \ref{tab:xg_hyper} displays the optimal hyperpa-
+rameters, providing insight into the XGBoost algorithm’s structure.
+Using gradient boosting, XGBoost sequentially builds trees to correct errors.
+(i) Features are sampled at 80\% per tree ('Colsample\_byTree = 0.8') to reduce overfitting.
+(ii) We put constraints on overall complexity with a maximum depth of 6 ('Max Depth = 6').
+(iii) Regularization terms, gamma (set to 0) and lambda (set to 0.1) help control overfitting.
+(iv) By employing a learning rate of 0.1 ('Learning Rate = 0.1'), each tree's contribution is shrunk to enhance generalization.
+(v) With 200 boosting rounds ('N Estimators = 200'), model performance iteratively improves.
+With this combination of hyperparameters, the algorithm should provide accurate results on test data while ensuring robustness.
 
-\emph{Accuracy Scores} The table below displays accuracy measures on the testing data using the best hyperparameters.
+\emph{Training Accuracy} The table below displays accuracy measures on the training data using 5 fold cross validation, along with the training time.
 
@@
 \centering
-\caption{Linear vs.\ Polynomial Performance on Testing Data}
+\caption{Random Forests vs.\ XGBoost on Training Data}
 \begin{tabular}{|c|c|c|c|}
 \hline
-Method & MSE & MAE & $R^2$ \\
+Method & Mean MSE & Standard Deviation & Training Time \\
 \hline
-Random Forests & 228,347.747 & 247.141 & 0.9803 \\
+Random Forests & 232,322.902 & 17741.65 & 0.453(s) \\
 \hline
-XGBoost & 195,937.71 & 233.941 & 0.9831 \\
+XGBoost & 171,332.011 & 18,084.45 & 0.254(s) \\
 \hline
 \end{tabular}
-\label{tab:mytable}
+\label{tab:dt_trn}
 \end{table}
 
-Comparing the accuracy scores, we see that both Random Forests and XGBoost perform well, with XGBoost having slightly lower MSE and MAE, and a slightly higher $R^2$, indicating better performance overall.
-However, the difference is relatively small, and both models seem to be capturing the underlying patterns in the data effectively.
-They both perform better than regression methosd.
+Table \ref{tab:dt_trn} displays the average training accuracy measures of Random Forests and XGBoost, alongside their respective training times.
+Notably, both algorithms achieved significantly lower training errors compared to the regression models.
+Random Forests exhibit a mean MSE of 232,322.902 with a standard deviation of 17,741.65, while XGBoost achieves a lower mean MSE of 171,332.011 with a standard deviation of 18,084.45.
+Additionally, XGBoost demonstrates a shorter training time of 0.254 seconds compared to 0.453 seconds for Random Forests.
+These results indicate that XGBoost outperforms Random Forests in terms of accuracy while also being computationally more efficient.
 
@@
 
-We used a 5-fold cross-validation procedure with an extensive Grid Search approach to find the optimal set of hyperparameters for our neural network. Here are the results:
+Again, we employed a 5-fold cross-validation procedure with an extensive Grid Search approach to identify the optimal set of hyperparameters for our neural network. Here are the results:
 
@@
 \end{tabular}
-\label{tab:mytable}
+\label{tab:nn_hyper}
 \end{table}
 
-The grid search method results in the architecture of our neural network model.
-Specifically, the neural network architecture seems to consist of a single hidden layer with 64 neurons, employing a learning rate of 0.1, RMSProp optimizer, and no dropout regularization.
-It underwent training for 250 epochs with a batch size of 256.
-Here, we present the training versus testing loss (MSE) across epochs during the training of our neural network model, along with its final performance on the testing data.
+The grid search method presents in the architecture of our neural network model.
+Specifically, the neural network architecture consists of a single hidden layer with 64 neurons, a learning rate of 0.1 for the model convergence, RMSProp optimizer, and no dropout regularization.
+Note that for this paper, our search was limited on developing an artificial neural network with one hidden layer, and one may choose to add additional layers and deepen this network further.
+The number of epochs was determined to be 250.
+To illustrate the training and testing accuracy versus epoch, refer to the figure below.
 
@@
     \includegraphics[width=0.8\linewidth]{1000.png} % Replace "example-image" with the filename of your image
-    \caption{MAE vs Degree of Polynomial Regression model}
-    \label{fig:boxplots}
+    \caption{Error (MSE) vs Number of Epochs: The figure depicts the training and testing errors as a function of the number of epochs (1000).
+Overfitting becomes apparent after Epoch 450, as shown by an increase in testing error while the training error remains low.}
+    \label{fig:trn-tst-1000}
 \end{figure}
 
-From our results of training the neural network across 1000 epochs, we observe that the model effectively learns the weights.
-Specifically, both the training and testing errors significantly reduce during the initial stages of training.
+We trained the neural network model using the optimal set of hyperparameters across 1000 epochs to learn the weights and observe when the model begins to overfit our data.
+Figure \ref{fig:trn-tst-1000} suggests that overfitting occurs after epoch 450.
+Initially, both the training and testing errors significantly decrease as the model weights are learned iteratively.
 However, around 500 epochs, our neural network model begins to overfit the data: the testing error starts to increase significantly while the training error continues to decrease.
-This indicates that we can use 250 epochs as our stopping point for neural network training to avoid overfitting.
+This suggests that early stopping training at around 250 epochs can help prevent overfitting.
 
@@
     \includegraphics[width=0.8\linewidth]{250.png} % Replace "example-image" with the filename of your image
-    \caption{MAE vs Degree of Polynomial Regression model}
-    \label{fig:boxplots}
+    \caption{Error (MSE) vs Number of Epochs: The figure depicts the training and testing errors as a function of the number of epochs (250).
+Both training and testing errors remain low across the 250 epochs.}
+    \label{fig:trn-test-250}
 \end{figure}
 
-As we can see, based on our neural network architecture, the training and testing loss appear to go down quite a bit once we finish implementing across epochs. Thus this approach is good [...]
+Thus, Figure \ref{fig:trn-test-250} displays that 250 epochs is a suitable choice for our application. We now present the training accuracy and time on this training data.
 
@@
 \centering
-\caption{Neural Network Performance on Testing Data}
-\begin{tabular}{|c|c|c|c|}
+\caption{Neural Network Performance on Training Data}
+\begin{tabular}{|c|c|c|}
 \hline
-Method & MSE & MAE & $R^2$ \\
+Method & MSE & Training Time \\
 \hline
-Neural Network & 382,237.13 & 272.47 & 0.9652  \\
+Neural Network & 198,675.7 & 25.3 (s)  \\
 \hline
@@
 
+Based on our training data and the given optimal set of hyperparameters, the neural network model performs quite well on the training data with an MSE of 198,675.7.
+This MSE ranks as the second best, following XGBoost, and displays better predictive performance as compared to the regression models.
+However, notably, this model incurred a significantly longer training time of 25 seconds as compared to the other models.
+Thus, we gain insight into the performance of this model on the original training data.
+Next, we will examine the model's performance on the test set and compare overall results among the models.
+
 \subsection{Comparison of models on testing data}
 
-The table below displays the performance of each algorithm we've taken from above on the testing data along with its BIC criteria:
+The table below displays the performance of each algorithm we've taken from above on the testing data:
 
@@
 \centering
-\caption{Comparison of Algorithms}
+\caption{Performance of Algorithms on Testing Data: MSE, MAE, $R^2$}
 \begin{tabular}{|c|c|c|c|c|}
@@
 \hline
-Random Forests & 228,347.747 & 247.141 & 0.9803  \\
+Random Forests & 210,230.72 & 246.879 & 0.9808  \\
 \hline
-XGBoost & 195,937.71 & 233.941 & 0.9831 \\
+XGBoost & 166,912.918 & 223.123 & 0.984 \\
 \hline
-Neural Networks & 382,237.13 & 272.47 & 0.9652 \\
+Neural Networks & 188,9781.85 & 271.621 & 0.8280 \\
 \hline
@@
 
-The comparison of algorithms for diamond pricing reveals that XGBoost performs the best, achieving the lowest MSE (195,937.71) and MAE (233.941), along with the highest $R^2$ (0.9831), indicating it provides the most accurate price predictions and best explains the variance in diamond prices.
-Random Forests also demonstrate strong performance, closely following XGBoost.
-Neural Networks and Polynomial Regression are effective but exhibit higher error rates and slightly lower $R^2$ values, making them less precise.
-Linear Regression, though straightforward and easy to interpret, performs the worst with the highest MSE and MAE, and the lowest $R^2$, suggesting it is the least reliable for predicting diamond prices.
-Therefore, ensemble methods like XGBoost and Random Forests are the most effective models for accurate diamond pricing, while Linear Regression is the least suitable.
+The performance of each algorithm on the testing data presents intriguing insights.
+Linear regression, which served as a baseline model, exhibits the highest MSE and MAE among all algorithms.
+This suggests the limitation of linear regression in capturing the complexity of relationships between diamond pricing and its characteristics.
+Polynomial regression performs better, significantly reducing the MSE and indicating its ability to capture non-linear relationships present in the data.
+Interestingly, despite having a higher training MSE, our polynomial regression model outperforms linear regression in terms of all three metrics on the testing data.
+This suggests that the inherent quadratic relation observed during the EDA phase in figure \ref{fig:scatter} might indeed be present.
+Furthermore, both linear and polynomial regression models have the shortest training times, which could be advantageous given limited computational resources.
+
+Both decision tree methods Random Forests and XGBoost demonstrate substantial improvements in performance, achieving significantly lower MSE and MAE values along with higher $R^2$ scores.
+This suggests that ensemble methods excel in capturing the intricate patterns in diamond pricing, likely due to their ability to model complex interactions between features.
+In particular, we can see that XGBoost stands out for its superior predictive accuracy and $R^2$ scores as compared to the other models.
+The training complexity of these models lies between that of linear/polynomial regression and neural networks, and they offer superior performance on the testing data.
+
+Comparatively, the neural network model yields low MSE and MAE, though its coupled with a slightly lower $R^2$ score compared to the ensemble methods.
+This indicates slightly suboptimal performance on the testing data as compared to those methods.
+Despite its potential to capture complex relationships, the neural network may require further optimization or a more sophisticated architecture to achieve better results in predicting diamond prices.
+Additionally, the training time for neural networks is the highest, which could lead to computation scarcity if the model becomes more complex or if more training data is added.
+
+Overall, Random Forests and XGBoost emerge as the top performers on the testing data, showcasing their effectiveness in accurately predicting diamond prices.
+These models, with their capability to handle non-linear relationships and capture intricate patterns, offer valuable insights into the factors influencing diamond pricing and can guide pricing strategies and market analyses effectively.
 
@@
 Our analysis demonstrated that advanced machine learning techniques, particularly Random Forests, XGBoost, and neural networks, significantly outperformed traditional linear regression models in terms of predictive accuracy.
-The exploratory data analysis revealed that carat is the most influential feature, with strong correlations to price and other dimensions.
 
+The exploratory data analysis revealed that carat is the most influential feature, showing strong correlations with price and other dimensions.
 Our findings suggest that leveraging these advanced models can provide jewelers and consumers with more accurate pricing, enhancing decision-making and market transparency.
-Future work can explore integrating additional features, refining model hyperparameters, and applying these models to other datasets to generalize our findings further.
 
-By adopting robust machine learning methodologies, the diamond industry can benefit from more precise pricing strategies, ultimately leading to improved market efficiency and customer satisfaction.
+Overall, Random Forests and XGBoost emerged as the top performers, showcasing their effectiveness in handling non-linear relationships and capturing intricate patterns in the data.
+The neural network model also showed promise, though it requires further optimization to reach its full potential.
+Linear and polynomial regression models, while simpler and faster to train, were less effective in capturing the complexities of diamond pricing.
+By adopting robust machine learning methodologies, the diamond industry can benefit from more precise pricing strategies, leading to improved market efficiency and customer satisfaction.
+
+\subsection{Potential for Future Efforts}
+
+Future research can build on these findings by incorporating a larger dataset to capture more diverse patterns and improve model generalization, conducting a more comprehensive grid search with a larger hyperparameter space to find optimal settings for models, particularly neural networks, and integrating additional features such as market trends, regional price variations, and other economic indicators to enhance model accuracy.
+Further optimization of neural network architectures, such as experimenting with deeper networks and different types of layers, could improve performance.
+Additionally, applying these models to other diamond datasets can validate their generalizability and robustness.
+By continuing to refine and expand upon the models and methodologies used in this study, the diamond industry can further enhance its pricing strategies and decision-making processes.
 
@@
 URL: https://www.cs.toronto.edu/~tijmen/csc321/slides/lecture\_slides\_lec6.pdf
-
 \end{thebibliography}

### 2024-06-08 paper/main.tex
added

@@
 Davis, United States \\
-XXX}
+acyeow@ucdavis.edu}
 \and
@@
 Davis, United States \\
-XXX}
+skodali@ucdavis.edu}
 \and
@@
 Davis, United States \\
-XXX}
+etoshniwal@ucdavis.edu}
 }

### 2024-06-08 paper/main.tex
Updates
fix tex file

--- 2024-06-04 paper/main.tex
@@
 \usepackage{xcolor}
+\usepackage{authblk}
 \bibliographystyle{plainnat}
@@
 
-\title{Machine Learning Approaches for Accurate Diamond Price Prediction\\
+\title{Machine Learning Based Approaches to Accurate Diamond Price Prediction\\
 \thanks{Final Project Report ECS-171: \textbf{Group 21}}
@@
 
-\author{
-\IEEEauthorblockN{Aditya Mittal}
-\IEEEauthorblockA{\textit{Department of Statistics} \\
-\textit{University of California, Davis}\\
-Davis, United States \\
-adimittal@ucdavis.edu}
-\and
-\IEEEauthorblockN{Andrew Yeow}
-\IEEEauthorblockA{\textit{Department of Computer Science} \\
-\textit{University of California, Davis}\\
-Davis, United States \\
-XXX}
-\and
-\IEEEauthorblockN{Yifan Cui}
-\IEEEauthorblockA{\textit{Department of Computer Science} \\
-\textit{University of California, Davis}\\
-Davis, United States \\
-harcui@ucdavis.edu}
-\and
-\IEEEauthorblockN{Nandini Koladi}
-\IEEEauthorblockA{\textit{Department of Computer Science} \\
-\textit{University of California, Davis}\\
-Davis, United States \\
-XXX}
-\and
-\IEEEauthorblockN{Eshan Toshniwal}
-\IEEEauthorblockA{\textit{Department of Computer Science} \\
-\textit{University of California, Davis}\\
-Davis, United States \\
-XXX}
-}
+% \author{
+% \IEEEauthorblockN{Aditya Mittal}
+% \IEEEauthorblockA{\textit{Department of Statistics} \\
+% \textit{University of California, Davis}\\
+% Davis, United States \\
+% adimittal@ucdavis.edu}
+% \and
+% \IEEEauthorblockN{Andrew Yeow}
+% \IEEEauthorblockA{\textit{Department of Computer Science} \\
+% \textit{University of California, Davis}\\
+% Davis, United States \\
+% XXX}
+% \and
+% \IEEEauthorblockN{Yifan Cui}
+% \IEEEauthorblockA{\textit{Department of Computer Science} \\
+% \textit{University of California, Davis}\\
+% Davis, United States \\
+% harcui@ucdavis.edu}
+% \and
+% \IEEEauthorblockN{Nandini Koladi}
+% \IEEEauthorblockA{\textit{Department of Computer Science} \\
+% \textit{University of California, Davis}\\
+% Davis, United States \\
+% XXX}
+% \and
+% \IEEEauthorblockN{Eshan Toshniwal}
+% \IEEEauthorblockA{\textit{Department of Computer Science} \\
+% \textit{University of California, Davis}\\
+% Davis, United States \\
+% XXX}
+% }
+
+\author[1]{Aditya Mittal}
+\author[2]{Andrew Yeow}
+\author[3]{Yifan Cui}
+\author[4]{Nandini Koladi}
+\author[5]{Eshan Toshniwal}
+\affil[1]{Department of Statistics; University of California, Davis}
+\affil[2]{Department of Computer Science; University of California, Davis}
+\affil[3]{Department of Computer Science; University of California, Davis}
+\affil[4]{Department of Computer Science; University of California, Davis}
+\affil[5]{Department of Computer Science; University of California, Davis}
 
@@
 In 2023, the U.S.
-jewelry market was valued at 73.32 billion USD, with diamonds comprising a significant share.
-Accurate diamond price prediction is essential for jewelers.
-This paper explores machine learning algorithms, including linear regression, polynomial regression, Random Forests, XGBoost, and neural networks, to predict diamond prices based on carat, cut, color, and clarity.
-Exploratory data analysis identified carat as the most influential feature.
-Our methodology involves rigorous model evaluation using metrics like Mean Squared Error (MSE), Mean Absolute Error (MAE), and R-squared (R²).
-Experimental results show XGBoost outperformed other models, achieving the lowest testing error, followed closely by Random Forests.
-Polynomial regression also performed well, highlighting the importance of capturing non-linear relationships.
-The neural network model, while promising, needs further optimization.
+jewelry market was valued at an approximate 73.32 billion USD, with diamonds capturing a significant share.
+Consequently, accurate diamond price prediction of diamonds is critical for both jewelers and consumers alike.
+This paper explores several machine learning algorithms, including regression models, tree-based ensemble methods, and artificial neural networks, to predict diamond prices based on important characteristics such as carat, cut, color, and clarity.
+Our exploratory data analysis identified carat as the most influential feature with regards to diamond prices.
+Our methodology involves rigorous model evaluation using metrics like Mean Squared Error (MSE), Mean Absolute Error (MAE), and R-squared ($R^2$) on testing data.
+The experimental results indicate that \emph{XGBoost} outperformed other models by achieving the lowest testing error, followed closely by Random Forests and the Artificial Neural Network.
+Polynomial regression also performed well, capturing the non-linear relationship between diamond prices and its features.
+The neural network model also demonstrated strong results for predictions, and will benefit from further tuning.
 By developing a robust predictive model, we aim to enhance pricing precision, benefiting both businesses and consumers in the diamond market.
@@
 \begin{IEEEkeywords}
-Diamond pricing, machine learning, predictive modeling, neural network, decision trees
+Diamond Pricing, Machine Learning, Artificial Neural Network, Ensemble Methods - Random Forests, XGBoost
 \end{IEEEkeywords}
@@
 As such, ensuring accurate price predictions becomes very important as it has significant financial implications on both businesses and consumers alike.
+Figure \ref{fig:diamond_intro} below demonstrates the importance of accurate diamond pricing.
+Two diamonds can differ significantly based on features such as carat, underscoring the need for precise price prediction methods.
+
+\begin{figure}[H]
+    \centering
+    \includegraphics[width=0.8\linewidth]{diamond_intro.png} % Replace "example-image" with the filename of your image
+    \caption{Two different diamonds differing in carat and overall pricing.
+The 1-carat diamond ranges from \$2,000- \$20,000, while the 2-carat diamond ranges from \$9,800 to \$64,500.
+This highlights the importance of accurate price prediction based on the given characteristics.}
+    \label{fig:diamond_intro}
+\end{figure}
 
@@
 Similarly, Mamonov and Triantoro \cite{Mamonov} studied the relationship between a diamond's physical attributes and respective prices in e-commerce contexts, with the goal of understanding how these attributes affect diamond prices.
-However, their study did not account for the diamond cut, a significant factor affecting market value, which is included in our dataset.
+However, their study did not account for the diamond cut – a significant factor affecting market value – which is indeed included in our dataset.
 Pandey et al.
@@
 Clearly, a variety of approaches have been taken to create accurate prediction models, ranging from investigating social contexts to individual characteristics.
-In this paper, we utilize several algorithms to create accurate models by building on the previous work.
+In this paper, we utilize several algorithms to create accurate models and build on this previous work.
 With this, the latter half of this section provides an overview of several machine learning algorithms applied in this paper, including regression models, decision trees, and neural networks.
@@
 
-With this, our literature review provides an overview of the research conducted on predicting diamond prices, including discussions on pertinent algorithms such as linear regression, decision trees, and neural networks.
+With this, our literature review provides a comprehensive overview of the previous efforts conducted on predicting diamond prices, including discussions on pertinent algorithms such as linear regression, decision trees, and neural networks.
 For hyperparameter tuning, we plan to utilize Grid Search to optimize the performance of both the decision trees and neural network models.
@@
     \centering
-    \includegraphics[width=0.53\linewidth]{Figure_1.png}
+    \includegraphics[width=0.60\linewidth]{distribution_plots.png}
     \caption{Distribution Plot of Continuous Features. Identify trends in our dataset such as skewness, location of outliers, mean, median, scale, etc.}
-    \label{fig:image_label}
+    \label{fig:distribution}
 \end{figure}
 
-The distribution of continuous predictors highlight potential skewness in our dataset.
+Figure \ref{fig:distribution} displays the distribution of continuous predictors and highlight potential skewness in our dataset.
 For instance, the predictor \emph{carat} suggests potential skewness to the right – with many upper outliers – as evidenced by its distribution plot.
@@
     \centering
-    \includegraphics[width=0.7\linewidth]{Figure_4.png} % Replace "example-image" with the filename of your image
-    \caption{Pair-wise Scatterplots for Each Feature vs. Diamond Price}
+    \includegraphics[width=0.61\linewidth]{scatter.png} 
+    \caption{Pairwise scatter-plots for Each Feature vs. Diamond Price. Displays a quadratic relationship between Carat, x, y, z and diamond pricing.}
     \label{fig:scatter}
@@
 
-From the scatter-plots of each feature against the response variable, we observe a quadratic relationship between the features, carat, x, y, and z, with respect to the the response variable price.
+Figure \ref{fig:scatter} displays the scatter-plots of each feature against the response variable, we observe a quadratic relationship between the features, carat, x, y, and z, with respect to the the response variable price.
 This suggests that a polynomial regression model might be suitable for our dataset in addition to a linear regression model.
@@
     \centering
-    \includegraphics[width=0.7\linewidth]{Figure_3.png} % Replace "example-image" with the filename of your image
-    \caption{Correlation Heat-Map for Continuous Features vs Diamond Price}
-    \label{fig:image_label}
+    \includegraphics[width=0.7\linewidth]{correlation.png} % Replace "example-image" with the filename of your image
+    \caption{Correlation Heat-Map for Continuous Features vs Diamond Price. It reveals that the feature 'carat' has the highest correlation (0.92) with diamond pricing.}
+    \label{fig:corr}
 \end{figure}
 
-Based on the results of our correlation heatmap, the feature \emph{Carat} appears to be highly correlated with price, with a correlation coefficient of 0.92.
+Based on the results of our correlation heatmap in figure \ref{fig:corr}, the feature \emph{Carat} appears to be highly correlated with price, with a correlation coefficient of 0.92.
 Similarly, it is correlated with dimension features \emph{x}, \emph{y}, and \emph{z}.
@@
     \centering
-    \includegraphics[width=0.8\linewidth]{figure_5.png} % Replace "example-image" with the filename of your image
-    \caption{Histogram for the distribution of categorical features: Cut, Clarity, and Color }
-    \label{fig:image_label}
+    \includegraphics[width=0.8\linewidth]{histogram.png} 
+    \caption{Histogram for the distribution of categorical features: Cut, Clarity, and Color.}
+    \label{fig:histogram}
 \end{figure}
@@
 This initial model will help us gain intuition about the relationship between the features and diamond prices.
-Additionally, since our Exploratory Data Analysis (EDA) results indicating a potential quadratic relationship between price and carat \& diamond dimensions (x, y, z), we also implement a Polynomial Regression model and test its performance.
-Using 5-fold cross-validation, we test powers ranging from 1 to 5 to determine the optimal degree of our regression model.
+Additionally, since our results in \ref{fig:scatter} suggested a potential quadratic relationship between price and carat \& diamond dimensions (x, y, z), we also implement a Polynomial Regression model and test its performance.
+Using 5-fold cross-validation, we test powers ranging from 1 to 3 to determine the optimal degree of our regression model.
 This allows us to identify whether more complex relationships in our data exist, capture nonlinear relationships, and assess whether Polynomial Regression outperforms traditional Linear Regression.
@@
 \centering
-\caption{Optimal Hyperparameter for Random Forests}
+\caption{Optimal Hyperparameter for Neural Networks}
 \begin{tabular}{|c|c|}
@@
 
-The grid search method presents in the architecture of our neural network model.
+The grid search method presents the architecture of our neural network model.
 Specifically, the neural network architecture consists of a single hidden layer with 64 neurons, a learning rate of 0.1 for the model convergence, RMSProp optimizer, and no dropout regularization.
@@
     \centering
-    \includegraphics[width=0.8\linewidth]{1000.png} % Replace "example-image" with the filename of your image
+    \includegraphics[width=1\linewidth]{1000.png} % Replace "example-image" with the filename of your image
     \caption{Error (MSE) vs Number of Epochs: The figure depicts the training and testing errors as a function of the number of epochs (1000).
@@
 Further optimization of neural network architectures, such as experimenting with deeper networks and different types of layers, could improve performance.
-Additionally, applying these models to other diamond datasets can validate their generalizability and robustness.
-By continuing to refine and expand upon the models and methodologies used in this study, the diamond industry can further enhance its pricing strategies and decision-making processes.
+Additionally, applying these models to other diamond datasets can validate their generalizability and overall robustness.
+We can continue to refine and expand upon the models and methodologies used in this study, and the diamond industry can further enhance its pricing strategies and decision-making processes.
 
@@
 
-Our project involves a regression-supervised learning task in the context of predicting the price of a diamond based on several different measures.
-Our dataset contains 53,940 observations and 9 columns (8 predictors and 1 response).
-The output variable “price” is continuous and has a range between \$326 to \$18,823.
-With this, our goal for this project is to create a machine-learning model using deep neural networks that can predict diamond prices as accurately as possible.
-The practical application of this project is quite significant, as an effective model can assist researchers and industry professionals in determining the value of diamonds in the market.
-As a part of developing our machine learning pipeline, we will first begin with exploratory data analysis.
-We do not have any missing values present, so we don’t have to worry about any data imputation for this specific project.
-First, we will begin by plotting the distribution of our features to understand the spread of our data and account for potential skewness.
-Since outliers can have significant effects on the overall accuracy of machine learning models, we will look for them in our dataset and remove them.
-We will use various outlier detection techniques such as one-class SVM and through the IQR computation; we're still planning the specific details as to how we will approach this step.
-Furthermore, since we have 3 categorical variables, we will use one-hot encoding and label encoding (in the case of an ordinal column) to convert them into numeric columns.
-Ultimately, our goal with EDA is to clean the dataset and understand the relationship between the predictors and the response variable.
-In terms of timeline, we plan to dedicate the EDA portion of our dataset this week/early next week (~ 1/1.5 weeks) before we begin training the model.
-We aim to have this portion of the project completed before the quarter midpoint and begin the process of training the model as well.
+To conclude, our project involves a regression-supervised learning task to predict the price of a diamond based on several different features.
+Our machine learning pipeline involved several steps: (i) \textbf{Exploratory Data Analysis (EDA):} During the first three weeks, our group focused on the exploratory data analysis portion.
+This took about 3-4 weeks.
+We began by plotting the distribution of our features to understand the spread of our data and account for potential skewness.
+Since outliers can significantly affect the overall accuracy of machine learning models, we used various outlier detection techniques, including the IQR computation, to identify and remove them.
+Our group contributed equally to this component: Andrew and Yifan focused on outlier detection, while Aditya, Eshan, and Nandini worked on creating the plots.
+(ii) \textbf{Model Implementation:} During the next three weeks, we implemented various models.
+Aditya worked on implementing the decision tree models and performed the necessary grid search for hyperparameter tuning.
+Andrew and Yifan focused on the neural network and regression models.
+This process took about three weeks (weeks 4-8 of the quarter).
+This step was the most rigorous and time consuming, as we evaluated each of the models against one another and tested many variations of hyperparameter spaces to find the optimal set in our application.
+During this time, Aditya also began writing the final report to ensure we stayed on schedule.
+(iii) \textbf{Report Writing:} The paper was written over the course of the quarter, once model training was completed to some extent.
+In weeks 9-10, the team finished the paper together.
+(iv) \textbf{Final Touches:} During the final week, we implemented finishing touches on the project report and integrated the model weights into the project GUI.
+Nandini and Eshan handled the HTML connections.
+Overall, our project demonstrated the importance of accurate diamond price prediction and showcased the effective use of machine learning models in achieving this goal.
 
-Once EDA is complete, we are going to use deep neural networks to train a machine learning algorithm that will most accurately predict diamond prices.
-Neural networks are quite complex and include many hyper-parameters, so we want to explore several of them and see how they can help improve our results.
-In particular, we will begin by using K-fold cross-validation to train several neural networks of different sizes (with the number of layers and neurons varying) and compare the performance of each method in predicting diamond price.
-We will use the metrics learned in class, using total error and MSE as benchmarks for overall model performance.
-Furthermore, we are interested in exploring regularization techniques such as L2 regularization (possibly dropout/weight-decay on the learning rate as well, if time permits) to reduce overfitting and further improve the total error.
-Our goal is to produce the best model in terms of accuracy while also achieving a low training time; thus we will explore the implementation of different optimization techniques such as momentum SGD/RMSProp/Adam (choose one of three, likely momentum GD since we will be able to explain the concepts behind it clearly) to see how they affect our results in comparison to gradient descent.
-The details and concept of our optimization techniques will be included as part of the literature review section in the project report.
-Lastly, we may explore the effect of different activation functions (sigmoid/tanh) to see how the model performs in terms of accuracy and performance when non-linearity is introduced.
-Ultimately, our final goal is to develop a model such that we can get the best prediction accuracy on the generalized unseen data.
-
-In conclusion, this project not only aims to give a precise price prediction but also to provide insight into what important factors influence diamond prices—specific factors that may be of high value to a consumer and seller of diamonds.
-Rigorous data analysis and model-tuning efforts will be put in place to realize a robust model that can help demystify the complexities behind diamond pricing.
-In terms of the specific methodologies learned in our class, this project will help us understand how changes in the neural network architecture and tuning of hyper-parameters can affect our overall results, including changes in key metrics such as model accuracy and overall training speed.
-
-\emph{Group Contributions} The group had equal contributions throughout the course of this project.
+\textbf{Group Contributions:} The group had equal contributions throughout the course of this project.
 Andrew, Yifan, and Aditya focused on the model training and writing the paper.
 Nandini and Eshan worked on exploratory data analysis, setting up the Github repo, and creating the Project GUI environment for users.
+Aditya remained in contact with the TA and professors to ensure the groups needs were met and all inquiries we resolved in a timely manner.
 

### 2024-06-08 code/project_code.py
Update project_code.py

@@ -25,7 +25,7 @@ import joblib
 
 ### Exploratory Data Analysis
 ### ---------------------------------------------------------------------------------
-df = pd.read_csv("~/Desktop/Year_three/Spring_2024/ECS171/ECS171_Final/diamonds.csv") 
+df = pd.read_csv("../dataset/diamonds.csv") 
 df = df.iloc[:, 1:]  # remove index column          
 # df.head()
 

### 2024-06-08 flask_gui/app.py
Update app.py

@@ -35,7 +35,7 @@ scaler = joblib.load('scaler.joblib')
 
 @app.route('/')
 def home():
-    return render_template('index.html')
+    return render_template('templates/index.html')
 
 @app.route('/result', methods=['POST'])
 def predict():
@@ -79,10 +79,10 @@ def predict():
             predictions['neural_network'] = "Model not available"
         
         # Return the predicted results to the user
-        return render_template("results.html", predictions=predictions)
+        return render_template("templates/results.html", predictions=predictions)
     except Exception as e:
         print("Error:", e)
-        return render_template("error.html")
+        return render_template("templates/error.html")
 
 if __name__ == '__main__':
     app.run(debug=True)

### 2024-06-08 flask_gui/app.py
fix app.py

@@ -35,7 +35,7 @@ scaler = joblib.load('scaler.joblib')
 
 @app.route('/')
 def home():
-    return render_template('templates/index.html')
+    return render_template('index.html')
 
 @app.route('/result', methods=['POST'])
 def predict():
@@ -54,13 +54,13 @@ def predict():
         # Prepare input data for prediction
         input_data = np.array([[carat, depth, table, x, y, z, cut_encoded, color_encoded, clarity_encoded]])
         
-        # Preprocess input data for polynomial regression model
-        poly = PolynomialFeatures(2)
-        input_data_poly = poly.fit_transform(input_data)
-        
         # Scale the input data
         input_data_scaled = scaler.transform(input_data)
         
+        # Preprocess input data for polynomial regression model
+        poly = PolynomialFeatures(2)
+        input_data_poly = poly.fit_transform(input_data_scaled)
+   
         # Make predictions using all models
         predictions = {}
         for model_name, model in models.items():
@@ -79,10 +79,10 @@ def predict():
             predictions['neural_network'] = "Model not available"
         
         # Return the predicted results to the user
-        return render_template("templates/results.html", predictions=predictions)
+        return render_template("results.html", predictions=predictions)
     except Exception as e:
         print("Error:", e)
-        return render_template("templates/error.html")
+        return render_template("error.html")
 
 if __name__ == '__main__':
     app.run(debug=True)

### 2024-06-08 other/eda.py
final changes

@@ -1,129 +1,0 @@
-# import necessary modules
-import numpy as np
-import pandas as pd
-import matplotlib.pyplot as plt
-import seaborn as sns
-import os
-
-from sklearn.linear_model import LinearRegression
-from sklearn.preprocessing import PolynomialFeatures
-from sklearn.preprocessing import StandardScaler
-from sklearn.model_selection import train_test_split
-from sklearn.metrics import mean_squared_error, r2_score
-
-import tensorflow as tf
-from tensorflow.keras.models import Sequential
-from tensorflow.keras.layers import Dense
-
-### DATA-Preprocessing & EDA Code
-
-# load data
-current_directory = os.path.dirname(os.path.abspath(__file__))
-parent_dir = os.path.dirname(current_directory)
-os.chdir(parent_dir)
-print(parent_dir)
-df = pd.read_csv("diamonds.csv") 
-df = df.iloc[:, 1:]               # remove index column
-# df.head()
-
-### ----------------------------------------------------------------------------
-
-### Plot distribution of each continous feature
-features_to_plot = ['carat', 'depth', 'table', 'x','y','z']
-plt.figure(figsize=(8, 6))
-for i, feature in enumerate(features_to_plot, 1):
-    plt.subplot(2, 3, i)
-    sns.histplot(df[feature], kde=True)
-    plt.title(feature)
-
-plt.tight_layout()
-plt.show()
-print("\nmean:\n{} , \nmedian:\n{} , \nstd:\n {}".format(df[features_to_plot].mean() , 
-                                                         df[features_to_plot].median(), 
-                                                         df[features_to_plot].std()))
-
-### ----------------------------------------------------------------------------
-
-### Frequency of categorical vars
-categorical_vars = ["cut", "color", "clarity"]
-fig, axes = plt.subplots(1, 3, figsize=(15, 5))
-for i, var in enumerate(categorical_vars):
-    sns.countplot(data=df, x=var, ax=axes[i])
-    axes[i].set_title("Frequency of " + var)
-    axes[i].set_xlabel(var)
-    axes[i].set_ylabel("Frequency")
-    axes[i].tick_params(axis='x', rotation=45)  
-plt.tight_layout()
-plt.show()
-
[…]
